

Here are some specific things that are needless hard in HOL Light.

\subsection{search}

Proof assistants should be less dependent on large technical
vocabularies.  The index to the HOL Light reference manual has $914$
tactics, conversions, rules, and other entries.  This does not include 
the identifiers of over
$11$ thousand theorems (with Flyspeck loaded).  
By contrast, a mathematician does not need to
know the official name of a single theorem to use it.  

There should be an entire search suite designed to make searching
through this easier.  It should take cues from programs such as
google, gmail, grep, find, quicksilver, Mac spotlight, and hoogle.

There are various hypergraphs that can be attached to the HOL system such
as $G_1$ with nodes the set of constants and a multiedge for each theorem
(containing the set of constants in the theorem).  The directed graph $G_2$
has nodes set equal to the set of theorems and directed edges if theorem $x$
is quoted in the proof of theorem $y$.  The search feature might do analysis
on these graphs in the style of google search.

There should be an extensible {\it suggest} command that is
context sensitive.  It takes the
current goal state and gives a menu of suggestions for what to try.  For example if
\verb!@! matches a subterm of the goal, it would suggest \verb!SELECT_ELIM_TAC!
or if the goal is a conjunction, it would suggest \verb!CONJ_TAC!

\subsection{parsing and printing}

The error messages for parsing.  If the term does not parse.
\begin{verbatim}
Exception: Noparse.
\end{verbatim}

If the preterm does not type check:
\begin{verbatim}
Exception:Failure"typechecking error (initial type assignment)".
\end{verbatim}

It would be helpful to know specifically which subterms do not typecheck.
This would involve catching the exception and processing it with further
diagnostics.

\subsection{assumption list support}

HOL Light can deal with the full assumption list or with the top assumption,
but it is hard to single out a  group of assumptions from the list.


\subsection{long proofs}

Long proofs are hard to deal with.

Need strategies for breaking longer proofs into shorter ones.

Example: \verb!countable_s! needs to accumulate several smaller
conclusions before reaching the major conclusion.  There are nine
subgoals.

\subsection{text editor support}

Most members of the group are using xpp.  In a survey on July 6, six
were using xpp and only two were using emacs.


Wish list.

Support for multiple sessions.  (I want to load a new sesssion while another
is still being used.)




