\documentclass[11pt]{amsart}

%Packages
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{alltt}
\usepackage{txfonts}
\usepackage{multind}
\usepackage{url}

% Macros
\input{flag}
\input{kepmacros}

% Theorem Stuff
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Assumption}
\newtheorem{corollary}{Corollary}
\newtheorem{remark}{Remark}

% Header
\title{Lemmas in Elementary Geometry\\(Unfinished Notes)}
\author{Thomas C. Hales}
\address{University of Pittsburgh, Pittsburgh PA}

% Index
\makeindex{guid}

%\makeindex

% Document
\begin{document}
\maketitle

\symbolfootnote[0]%
 {This research has been supported by NSF grants 0503447 and 0804189.\hfil\break}
\symbolfootnote[0]{\hfil\break}
\symbolfootnote[0]{A recent pdf version of the document is maintained at \url{http://www.math.pitt.edu/~thales/papers/Lemmas_Elementary_Geometry.pdf}\hfil\break}
\footnote{The work on this collection has stopped, at least for now, because the author is currently pursuing an alternate approach to the proof of the Kepler conjecture, based on C. Marchal's objective function.  The original purpose of this collection of problems was a step in the Flyspeck project.  If Marchal's approach can be pushed to completion, then this collection no longer serves a purpose in the Flyspeck project.  The document still has a few errors (hopefully all minor).   -T. Hales, April 10, 2009} %XX 

 
% Body


\input{tarski}



% Endmatter


\newpage
\begin{thebibliography}{CHMS94}
\bibitem[Hal97a]{part1} Thomas C. Hales, Sphere Packings I,
    Discrete and Computational Geometry, 17 (1997), 1-51.
\bibitem[DCG]{DCG} Thomas C. Hales, The Kepler Conjecture, Discrete and
Computational Geometry, July 2006. (There are several papers in
the issue, but the section numbers run through the entire issue.)
\bibitem[Hal07b]{EZ} Thomas C. Hales, Easy Pieces in Geometry, preprint, 2007.
\bibitem[Hal07c]{MG} Thomas C. Hales, Some Methods of Problem
Solving in Geometry, Proc. of Logic in Computer Science, IEEE, 2007.
\end{thebibliography}

{

%\twocolumn{
%\input{guid.ind}}

}

% Index
%\newpage
%\printindex
\tableofcontents
\end{document}

