% Document created Dec 6, 2002
% Composite paper of the Kepler conjecture
% as requested by the Annals of Mathematics
% Author Thomas C. Hales
% Copyright Thomas C. Hales

% Revised March 10, 2005.  Sam's thesis added.
% Revised July 19, 2005. Sam's revisions added.  (See changes_2005_7_19.txt)
% Checked that latex files still work (using XeLaTeX, October 2015)

\documentclass[onecolumn]{newsiambook}

\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{latexsym}
\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{amsmath}
%\usepackage{mathidx}
\usepackage{makeidx}
\usepackage{multicol}
\usepackage{crop}

 \crop
 \makeindex

%% TAG ANNALS VERSION SECTIONS
\def\noversion#1{}
    %\iftrue % shortversion
    \iffalse % longversion

        % ANNALS VERSION HERE.
        \def\shortversion#1{#1}
        \def\longversion#1{}
        \def\chap{chapter{}}
        \def\Chap{Chapter{}}
        \def\Chaps{Chapters{}}
        \def\chaps{chapters{}}
        \def\paper{paper{}}
    \else
        % DCG UNABRIDGED DEFS HERE.
        \def\shortversion#1{}
        \def\longversion#1{#1}
        \def\Part{Paper{}}
        \def\chap{section{}}
        \def\Chap{Section{}}
        \def\Chaps{Sections{}}
        \def\chaps{sections{}}
        \def\paper{volume{}}
        \renewcommand\chaptername{Section}
        \renewcommand\partname{Paper}
    \fi

% NOTES:
% 1. Give exact citation on algorithms.
% 2. Put archival material at LANL, with better title, etc.

%% non conventions:
%% non-Euclidean, non-triangular, non-strict, non-star, non-aggregated
%% rest omit the hyphen: nonoverlapping nonlinear, nonzero, nonpositive, nonnegative,
%%    nonempty, nonadjacent, nondegenerate

%XX
% 1. regenerate index.
% 2. def style -- italic
% 4. Skipped 8.4 and 8.5
% 6. links on equations.
% 7. hyperlinks.

% PDF LATEX version.
% must first convert all graphics to .jpg format.
%\usepackage{hyperref} % for pdflatex version.


\include{kepmacros}

% MACROS THAT SHOULD NOT APPEAR IN INDIVIDUAL TEMPLATES
% SIAM BOOK CLASS SPECIFIC
\newtheorem{lem}{Lemma}[chapter]
\newtheorem{thm}{Theorem}[chapter]
\newtheorem{prop}{Proposition}[chapter]
\newtheorem{cor}{Corollary}[chapter]
\newtheorem{defn}{Definition}[chapter]

\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{calculation}[theorem]{Calculation}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{observation}[theorem]{Observation}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
    %% Title, optional running title in brackets.
    \title{A Proof of the Kepler Conjecture
      \longversion{(DCG Version)}
      \shortversion{(Annals Version)}
        }
    %\authorline{Thomas C. Hales}
    \author{Thomas C. Hales}
    \maketitle
    \frontmatter
    \tableofcontents
    \thanks

\begin{thepreface}

This project would not have been possible without the generous
support of many people.   I would particularly like to thank Kerri
Smith, Sam Ferguson, Sean McLaughlin,  Jeff Lagarias, Gabor Fejes
T\'oth, Robert MacPherson, and the referees for their support of
this project.  A more comprehensive list of those who contributed
to this project in various ways appears in
\shortversion{\cite{historical}}
\longversion{\Chap~\ref{sec:ack}}.


\bigskip

This research was supported by a grant from the NSF over the
period 1995--1998.

\bigskip

Version -- July, 2005.

\bigskip


\smallskip
\longversion{\newpage}

%% abstracts:
\def\abst#1#2{\longversion{\noindent{\bf Abstract: #1}\ \ \input{#2}\bigskip}}
\abst{Historical Overview of the Kepler Conjecture}{abstract1}

\abst{A Formulation of the Kepler Conjecture}{abstract2}

\abst{Sphere Packings III. Extremal Cases}{abstract3}

\abst{Sphere Packings IV. Detailed Bounds}{abstract4}

\abst{Sphere Packings V. Pentahedral Prisms}{abstract5}

\abst{Sphere Packings VI. Tame Graphs and Linear Programs}
    {abstract6}

\end{thepreface}

\mainmatter


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\longversion{
   \part{Historical Overview of the Kepler Conjecture}
   \label{part:intro}
   \chapter{Introduction}
   \input{intro1}
   \label{sec:intro-review}
   \input{historical}
    }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\longversion{
    \part{A Formulation of the Kepler Conjecture}
    \input{intro2}
    }
    \label{part:form}
    \input{form}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\longversion{
    \part{Sphere Packings III. Extremal Cases}
    \label{part:iii}
    \input{intro3}
    \input{local_opt}
    \input{fine}
    }
\shortversion{\input{local_opt}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\longversion{
    \part{Sphere Packings IV. Detailed Bounds}
    \label{part:bounds}
    \input{intro4}
    \input{part4}
    }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\longversion{
    \part{Sphere Packings V. Pentahedral Prisms -- Ferguson}
    \label{part:ferguson}
    \chapter*{Introduction}
    \input{intro5}
    \input{ferguson}
    }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\longversion{
    \part{Sphere Packings VI. Tame Graphs and Linear Programs}
    \label{part:tame}
    \input{intro6}
    \input{tame}
    \input{linprog}
    \input{implement}
    }

\shortversion{
    \label{part:tame}
    \input{tame}
    \input{linprog}
    }

\section{Conclusion}

By combinations of branching along the lines set forth in the
preceding sections, a sequence of linear programs is obtained that
establishes that  $\sigma(D)$ is less than $8\,\pt$.  For details
of particular cases, the interested reader can consult the log
files in \cite{web}, which record which branches are followed for
any given tame graph.  (For most tame graphs, a single linear
program suffices.)

\smallskip

This completes the \shortversion{(abridged)} proof of the Kepler
conjecture.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\backmatter

\input{kep_bib}

\printindex

\end{document}
