%% REFERENCES FOR HISTORICAL OVERVIEW

\bibitem[arXiv]{arXiv} http://xxx.lanl.gov.

\bibitem[Ben74]{Ben74}  Bender, C., Bestimmung der gr\"ossten Anzahl gleich
grosser Kugeln, welche sich auf eine Kugel von demselben Radius, wie
die \"ubrigen, auflegen lassen, {\it Archiv Math. Physik} 56 (1874),
302--306.

\bibitem[Bez90]{Bez90} A. Bezdek and W. Kuperberg, Maximum density space packing with
    congruent circular cylinders of infinite length,
    {\it Mathematica} 37 (1990), 74--80.

\bibitem[BKM91]{BKM91} A. Bezdek, W. Kuperberg, and E. Makai Jr., Maximum density
    space packing with parallel strings of balls,
    {\it DCG} 6 (1991), 227--283.

\bibitem[Bez94]{Bez94} A. Bezdek, A remark on the packing density in the 3-space
    in {\it Intuitive Geometry}, ed. K. B\"or\"oczky and G. Fejes
    T\'oth, {\it Colloquia Math. Soc. J\'anos Bolyai} 63, North-Holland
    (1994), 17--22.

\bibitem[Bez97]{Bez97} K. Bezdek, Isoperimetric inequalities and the dodecahedral
    conjecture, {\it Internat. J. Math.} 8, no. 6 (1997), 759--780.

\bibitem[Bli19]{Bli19} H. F. Blichfeldt,
    Report on the theory of the geometry of numbers,
    {\it Bull. AMS}, 25 (1919), 449--453.

\bibitem[Bli29]{Bli29} H. F. Blichfeldt,
    The minimum value of quadratic forms and the closest
    packing of spheres, {\it Math. Annalen} 101 (1929), 605--608.

\bibitem[Bli35]{Bli35} H. F. Blichfeldt,
    The minimum values of positive quadratic forms in six,
    seven and eight variables, {\it Math. Zeit.} 39 (1935), 1--15.

\bibitem[Boe52]{Boe52} Boerdijk, A. H. Some remarks concerning close-packing
of equal Spheres, {\it Philips Res. Rep.} 7 (1952), 303--313.

\bibitem[CK04]{CoKu} H. Cohn, A. Kumar, The densest lattice in twenty-four
dimensions, math.MG/0408174, (2004).

\bibitem[CHMS94]{CoHMS94} J. H. Conway, T. C. Hales, D. J. Muder, and N. J. A. Sloane,
    On the Kepler conjecture, {\it Math. Intelligencer} 16,
    no. 2 (1994), 5.

\bibitem[CS95]{CoSl95}  J. H. Conway,  N. J. A. Sloane, What are all the
best sphere packings in low dimensions? {\it DCG} 13 (1995),
383--403.

\bibitem[CS93]{CoSl93} J. H. Conway and N. J. A. Sloane, Sphere packings, lattices
    and groups,  third edition, Springer-Verlag, New York,
    1998.

\bibitem[Fej93]{Fej93} G. Fejes T\'oth and W. Kuperberg, Recent results in the
    theory of packing and covering, in New trends in
    discrete and computational geometry, ed. J. Pach, Springer
    1993, 251--279.

\bibitem[Fej95]{Fej95} G. Fejes T\'oth, Review of [Hsi93], {\it Math. Review} 95g\#52032, 1995.

\bibitem[Fej95b]{Fej95b} G. Fejes T\'oth, Densest packings of typical convex sets
    are not lattice-like, {\it DCG}, 14 (1995), 1--8.

\bibitem[Fej97]{Fej97} G. Fejes T\'oth, Recent progress on packing and covering,
     Advances
in Discrete and Computational Geometry, (South Hadley, MA, 1996),
pp. 145-162. Contemp. Math. 223 (1999), AMS, Providence, RI, 1999.
MR 99g:52036


\bibitem[Fej72]{Fej72} L. Fejes T\'oth, {\it Lagerungen in der Ebene auf der
    Kugel und im Raum}, second edition,
    Springer-Verlag, Berlin New York, 1972.

\bibitem[Fej64]{Fej64} L. Fejes T\'oth, Regular figures, Pergamon Press,
    Oxford London New York, 1964.

\bibitem[Fej42]{Fej42} L. Fejes T\'oth,  \"Uber die dichteste Kugellagerung,
{\it Math. Zeit.} 48 (1942 1943), 676--684.

\bibitem[Fej50]{Fej50} L. Fejes T\'oth, Some packing and covering theorems,
    {\it Acta Scientiarum Mathematicarum (Szeded)} 12/A, 62--67.

\bibitem[Fej53]{Fej53} L. Fejes T\'oth, {\it Lagerungen in der Ebene auf
der Kugel und im Raum}, Springer, Berlin, first edition, 1953.

\bibitem[FH98]{Form} S. P. Ferguson and T. C. Hales, A formulation of
the Kepler Conjecture, preprint 1998.

\bibitem[Fer97]{Fer97} S. P. Ferguson, Sphere Packings V, thesis, University of
Michigan, 1997.

\bibitem[Gau31]{Gau31} C. F. Gauss, Untersuchungen \"uber die Eigenscahften der
positiven tern\"aren quadratischen Formen von Ludwig August Seber,
    {\it G\"ottingische gelehrte Anzeigen}, 1831 Juli 9,
also published in {\it J. reine angew. Math.} 20 (1840), 312--320,
and
    {\it Werke},  vol. 2,
    K\"onigliche Gesellschaft der Wissenschaften, G\"ottingen,
            1876, 188--196.

\bibitem[Goo97]{Goo97} J. E. Goodman and J. O'Rourke, Handbook of discrete and
    computational geometry, CRC, Boca Raton and New York, 1997.

\bibitem[Gun75]{Gun75} S. G\"unther, {\it Ein stereometrisches Problem},
{\it Archiv der Math. Physik} 57 (1875), 209--215.


\bibitem[Hal92]{Hal92} T. C. Hales, The sphere packing problem, {\it J. Computational
    Applied Math.} 44 (1992), 41--76.

\bibitem[Hal93]{Hal93} T. C. Hales, Remarks on the density of sphere packings in
    three dimensions, {\it Combinatorica} 13 (1993), 181--187.

\bibitem[Hal94]{Hal94} T. C. Hales, The status of the Kepler conjecture,
    {\it Math. Intelligencer} 16, no. 3, (1994), 47--58.

\bibitem[Hal96]{Hal96} T. C. Hales,
    {http://www.pitt.edu/\~\relax thales/kepler98/holyoke.html}

\bibitem[Hal97a]{Hal97A} T. C. Hales, Sphere Packings I, Disc. Comp. Geom 17:1-51
(1977).

\bibitem[Hal97b]{Hal97B} T. C. Hales, Sphere Packings II, Disc, Comp. Geom
18:135--149 (1997).

\bibitem[Hal98a]{Hal98A} T. C. Hales, \hfill{http://www.math.pitt.edu/\~%
    \relax thales/kepler98/packings.html}.  \hfill\break
    The computer code is
    permanently archived at\hfill\break
    http://xxx.lanl.gov/abs/math.MG/9811078.

\bibitem[Hal98b]{Hal98B} T. C. Hales, Sphere Packings III, math.MG/9811075.

\bibitem[Hal98c]{Hal98C} T. C. Hales, Sphere Packings IV, math.MG/9811076.

\bibitem[Hal98d]{Hal98D} T. C. Hales, The Kepler Conjecture, math.MG/9811078.

\bibitem[Hal00]{Hal00} T. C. Hales, Cannonballs and Honeycombs, Notices of the
AMS, Vol 47, No. 4.

\bibitem[Hal01]{Hal01} T. C. Hales, Sphere Packings in 3 Dimensions,
Arbeitstagung, 2001, math.MG/0205208.

\bibitem[Hal03]{Hal-A} Thomas C. Hales, Some algorithms arising in the proof of
the Kepler Conjecture, Discrete and Computational Geometry: The
Goodman-Pollack Festschrift, Jacob E. Goodman (Edt), Springer
Verlag, July 2003.


\bibitem[Hil01]{Hil01} D. Hilbert, Mathematische Probleme, {\it Archiv Math. Physik} 1 (1901),
    44--63, also in {\it Proc. Sym. Pure Math.} 28 (1976), 1--34.

\bibitem[Hop74]{Hop74} Hoppe R. {\it Bemerkung der Redaction}, Math. Physik 56
(1874), 307-312.

\bibitem[HPT95]{HoPT95} R. Horst, P.M. Pardalos, N.V. Thoai, {\it Introduction
    to Global Optimization}, Kluwer, 1995.

\bibitem[Hsi93a]{Hsi93} W.-Y. Hsiang, On the sphere packing problem and the proof
    of Kepler's conjecture, Internat. J. Math 93 (1993), 739-831.

\bibitem[Hsi93b]{Hsi93a} W.-Y. Hsiang, On the sphere packing problem and the
    proof of Kepler's conjecture, in {\it Differential geometry and
    topology} (Alghero, 1992), World Scientific, River Edge,
    NJ, 1993,  117--127.

\bibitem[Hsi93c]{Hsi93b} W.-Y. Hsiang, The geometry of spheres, in {\it Differential
    geometry} (Shanghai, 1991), World Scientific, River Edge, NJ,
    1993, 92-107.

\bibitem[Hsi95]{Hsi95} W.-Y. Hsiang, A rejoinder to T. C. Hales's article ``The status
    of the Kepler conjecture,'' {\it Math. Intelligencer} 17, no. 1, (1995),
    35--42.

\bibitem[Hsi02]{Hsi02} W.-Y. Hsiang, Least Action Principle of Crystal Formation
of Dense Packing Type and the Proof of Kepler's Conjecture, World
Scientific, 2002.

\bibitem[Kar66]{Kar66} R. Kargon, Atomism in England from Hariot to Newton,
    Oxford, 1966.

\bibitem[Kep66]{Kep66} J. Kepler, The Six-cornered snowflake, Oxford Clarendon Press,
    Oxford, 1966,  forward by L. L. Whyte.

\bibitem[KZ73]{KoZ73} A. Korkine and  G. Zolotareff, Sur les formes quadratiques,
    {\it Math. Annalen} 6 (1873), 366--389.

\bibitem[KZ77]{KoZ77} A. Korkine and  G. Zolotareff, Sur les formes quadratiques
    positives, {\it Math. Annalen} 11 (1877), 242--292.

\bibitem[Lag73]{Lag73} J. L. Lagrange,  Recherches d'arithm\'etique, {\it Nov. Mem.
    Acad. Roy. Sc. Bell Lettres Berlin} 1773, in {\it \OE uvres}, vol. 3,
    693--758.

\bibitem[Lee56]{Lee56} J. Leech, The Problem of the Thirteen Spheres,
{\it The Mathematical Gazette}, Feb 1956, 22--23.

\bibitem[Lin86]{Lin86} J. H. Lindsey II, Sphere packing in $R^3$, {\it Mathematika}
    33 (1986), 137--147.

\bibitem[Mas66]{Mas66} B. J. Mason, On the shapes of snow crystals, in \cite{Kep66}.

\bibitem[McL98]{McL98} S. McLaughlin, A proof of the dodecahedral conjecture,
    preprint, math.MG/9811079.

\bibitem[Mel97]{Mel97} J. B. M. Melissen, Packing and covering with circles,
    Ph.D. dissertation, Univ. Utrecht, Dec. 1997.

\bibitem[Mil76]{Mil76} J. Milnor, Hilbert's problem 18: on crystallographic groups,
    fundamental domains, and on sphere packings, in
    Mathematical developments arising from Hilbert problems,
    {\it Proc. Symp. Pure Math.}, vol 28, 491--506, AMS, 1976.

\bibitem[MP93]{MP93} W. Moser, J. Pach, Research problems in discrete geometry,
    DIMACS Technical Report, 93032, 1993.

\bibitem[Mud88]{Mud88} D. J. Muder, Putting the best face on a Voronoi polyhedron,
    {\it Proc. London Math. Soc.} (3) 56 (1988), 329--348.


\bibitem[Mud93]{Mud93}  D. J. Muder A New Bound on the Local Density
of Sphere Packings, {\it Discrete and Comp. Geom.} 10 (1993),
351--375.

\bibitem[Mud97]{Mud97}  D. J. Muder, letter, in {\it Fermat's enigma}, by S. Singh,
        Walker, New York, 1997.

\bibitem[Oes90]{Oes90} J. Oesterl\'e,  Empilements de sph\`eres,
    S\'eminaire Bourbaki, vol. 1989/90, Ast\'erisque (1990),
        No. 189--190 exp. no. 727, 375--397.

\bibitem[PA95]{PaA95} J. Pach, P.K. Agarwal, {\it Combinatorial geometry}, John Wiley,
    New York 1995.

\bibitem[Plo00]{Plo00}  K. Plofker, private communication, January 2000.

\bibitem[Ran47]{Ran47} R. A. Rankin, {\it Annals of Math.} 48 (1947), 228--229.

\bibitem[Rog58]{Rog58} C. A. Rogers, The packing of equal spheres, {\it Proc. London Math.
    Soc.} (3) 8 (1958), 609--620.

\bibitem[Rog64]{Rog64} C. A. Rogers, {\it Packing and covering}, Cambridge University Press,
    Cambridge, 1964.

\bibitem[SW53]{Sch53} K. Sch\"utte and B.L. van der Waerden, Das
Problem der dreizehn Kugeln, {\it Math. Annalen} 125, (1953),
325--334.

\bibitem[SM44]{SeM44} B. Segre and K. Mahler, On the densest packing of
    circles, {\it Amer. Math Monthly} (1944), 261--270.

\bibitem[Shi83]{Shi83} J. W. Shirley,
{\it Thomas Harriot: a biography}, Oxford, 1983.

\bibitem[SHDC95]{SHDC95} N. J. A. Sloane, R. H. Hardin, T. D. S. Duff, J. H. Conway,
    Minimal-energy clusters of hard spheres,
    {\it DCG} 14,  no. 3, (1995), 237--259.

\bibitem[Szp02]{Szp02} G. G. Szpiro, Kepler's Conjecture, Wiley, 2002.

\bibitem[Thu92]{Thu92} A. Thue, Om nogle geometrisk taltheoretiske Theoremer,
    {\it Forandlingerneved de Skandinaviske Naturforskeres} 14 (1892), 352--353.

\bibitem[Thu10]{Thu10} A. Thue, \"Uber die dichteste Zusammenstellung von
    kongruenten Kreisen in der Ebene, {\it Christinia Vid. Selsk. Skr.} 1
    (1910), 1--9.

\bibitem[Why66]{Why66} L. L. Whyte, forward to \cite{Kep66}.
