%% OLD REFERENCES

\begin{thebibliography}{TCH97a}

\bibitem[CS98]{CS} J. H. Conway and N. J. A. Sloane, Sphere packings, lattices
    and groups,  third edition, Springer-Verlag, New York, 1998.

\bibitem[Fej72]{FT} L. Fejes T\'oth, {\it Lagerungen in der Ebene auf der
    Kugel und im Raum}, second edition,
    Springer-Verlag, Berlin New York, 1972.

\shortversion{\bibitem[Fer97]{thesis} Samuel P. Ferguson, Sphere
Packings V, thesis, University of Michigan,
    1997;  arXiv math.MG/9811077; to appear in Discrete and Computational
    Geometry.\index{Ferguson}}

\bibitem[Hal92]{spp} Thomas C. Hales, The Sphere Packing Problem, J. of Comp.
and App. Math. 44 (1992) 41--76.

\bibitem[Hal93]{remarks} Thomas C. Hales, Remarks on the Density of Sphere Packings,
        Combinatorica, 13 (2) (1993) 181-197.

\bibitem[Hal96]{reform} Thomas C. Hales, A reformulation of the
Kepler Conjecture, unpublished manuscript, Nov. 1996.

\bibitem[Hal97a]{part1} Thomas C. Hales, Sphere Packings I,
    Discrete and Computational Geometry, 17 (1997), 1-51.

\bibitem[Hal97b]{part2} Thomas C. Hales, Sphere Packings II,
    Discrete and Computational Geometry, 18 (1997), 135-149.

\bibitem[Hal00]{CH} Thomas C. Hales, Cannonballs and Honeycombs,
Notices Amer. Math. Soc.  47  (2000),  no. 4, 440--449.

\bibitem[Hal01]{arbeitstagung} Thomas C. Hales, Sphere Packings in 3
Dimensions, Arbeitstagung, 2001.

\bibitem[Hal03]{algorithm} Thomas C. Hales, Some algorithms arising in
the proof of the Kepler Conjecture,
 Discrete and computational geometry,
 489--507, Algorithms Combin., 25, Springer, Berlin, 2003.

\shortversion{\bibitem[Hal06a]{KC} Thomas C. Hales, A Proof of the
Kepler Conjecture (unabridged version), to appear in Discrete and
Computational Geometry.}

\longversion{\bibitem[Hal05a]{KC} Thomas C. Hales, A Proof of the
Kepler Conjecture (summary version), to appear in Annals of
Mathematics.}

\shortversion{\bibitem[Hal06b]{historical} Thomas C. Hales, An
Overview of the Kepler Conjecture, preprint, in \cite{KC}.}

\bibitem[Hal05b]{web} Thomas C. Hales, Packings, \hfill\break
    \hfill{\it http://www.math.pitt.edu/\~%
    \relax thales/kepler98.html} \hfil\break
     (The source code, inequalities,
    and other computer data relating to the solution is also found
    at {\it http://xxx.lanl.gov/abs/math/9811078v1}.)

\bibitem[IEEE]{IEEE} IEEE Standard for Binary Floating-Point
Arithmetic, ANSI/IEEE Std. 754-1985, IEEE, New York.

\bibitem[Hil01]{hilbert}
D. Hilbert, Mathematische Probleme, {\it Archiv Math. Physik} 1
(1901),
    44--63, also in {\it Proc. Sym. Pure Math.} 28 (1976), 1--34.

\bibitem[Rog58]{Rogers}
  C. A. Rogers, The packing of equal spheres, {\it Proc. London
Math.
    Soc.} (3) 8 (1958), 609--620.

\end{thebibliography}
