

\section*{Acknowledgments}

This book is just a blueprint.  It is a piece of paper that gives instructions
about how to construct a formal proof.  The actual construction of the edifice
is a large team effort.  My contributions have been a tiny fraction of the whole.

I am grateful to Sam Ferguson for the years that
he spent working on this problem with me.  I also thank the early editors
Robert MacPherson, Gabor Fejes T\'oth, and Jeff Lagarias for the many improvements
they brought to the original proof.

Many have worked tirelessly to make the Flyspeck formalization project
a reality.  I wish to thank Nguyen Quang
Truong, % fix accents. (he found a mistake)
Nguyen Tat Thang, % fix accents. (he found a mistake),
Tran Nam Trung, Trieu Thi Diep, Dang Tat Dat, Nguyen Duc Tam, Vu Quang
Thanh, Nguyen Duc Thinh, Hoang Le Truong, Ta Thi Hoai An, Erin Susick,
Laurel Beth Martin, Mary Johnston, Catalin Anghel, Jeremy Avigad,
Mark Adams, Freek Wiedijk, Tobias Nipkow, John Harrison, Sean
McLaughlin, Gertrud Bauer, Matthew Wampler-Doty, Steven Obua, and
Roland Zumkeller.
%Tran Nam Trung for many helpful
%comments, discussions, and corrections about the material in the
%chapter on hypermaps.  

 I would like to thank John Harrison for designing and implementing the most amazing piece of
 software I have ever encountered.  He has contributed large bodies of formal mathematics to
the libraries of the HOL Light proof assistant in order to make the Flyspeck project
possible.   Tobias Nipkow became involved in the project at an early stage and
directed the graduate work of two students, Bauer and Obua, on the formal verification of computer
code for the Flyspeck project. Mark Adams has taught me about how
 large-scale project should be organized and has skillfully managed 
  Flyspeck group in Hanoi.

  When I learned that my NSF proposal had been funded, I felt
  nauseated.  I considerd turning down the
  funding because I worried that the  deliverables I had put
  into the proposal were far beyond my capacity.  It was one thing to
  grandstand and quite another to implement the largest
  formalization project that had ever been attempted.  In retrospect, I
  am indebted to NSF for its generous support of this project.  It has supported
  this project in all of its aspects, including a
%Through this
%  research support, many people have contributed years of their time to move this
%  project forward.  A 
  two-month  workshop on formal proof
  and the Flyspeck project  at the Math Institute in Hanoi during the summer of
  2009.  This workshop and all that has grown out of it have been crucial for
  the success of this project.

I am also indebted to the Benter Foundation for a generous grant to rework the original
proof along lines proposed by Marchal.

Much of the material from this book was covered
in a course on discrete geometry and computers at the University of
Pittsburgh and then later %at a workshop on formal proof and the Flyspeck project 
in Hanoi.  I would like to thank the members of these
groups for assisting in the preparation of the book.

The nth draft of this book was written during a sabbatical leave from the
University of Pittsburgh.  I wish to thank the many institutions that
supported me during this period: the Max Planck Institute in Bonn, the
\'Ecole Normale Sup\'erieure in Paris, the Institute of Math in Hanoi, Radboud
University in Nijmegen, and the University of Strasbourg.  These visits would not
have been possible without the assistance of  Grigori Mints, Fran\c{c}ois Loeser,
Florence Lecomte, Henk Barendregt, Ha Huy Khoai and Ng\^o Vi\d{\^e}t Trung.

\bigskip

This book is dedicated to the memory of my grandfather Wayne B. Hales.
%who encouraged my childhood interest in the Bohr model of the atom.  I
%naturally wondered as an eleven-year-old boy how spherical electrons might
%be packed into each shell.

