% ------------------------------------------------------------ 
% Author: Thomas C. Hales 
% Format: LaTeX Book Chapter: Dense Sphere Packings
% ------------------------------------------------------------

\chapter{Credits}\label{sec:credit}

This book is just a blueprint, which gives instructions about how to
construct the edifice of a formal proof. 

\newterm{Flyspeck} is the name of an ongoing project to construct a
formal proof of the Kepler conjecture in the HOL Light proof
assistant, along the lines described in this book.  The eventual aim
of the project is to give a formal verification of the computer
portions of the proof as well as the standard text portions of the
proof.  The project is about 80\% complete as of May 2012.  Work will
continue until completion.  The source code for the project and
information about the current project status are available at
\cite{website:FlyspeckProject}.



Here is the fine print about the current project status. (I hope that
this status report is out of date by the time this book is printed.)
There are four components to the formalization project (text, hypermaps, linear programs,
and nonlinear inequalities), at various stages of completion.
\begin{enumerate}
\item
The English mathematical text in this book has been fully
  formalized, with the following explicit omissions.
\begin{enumerate}
\item The formal proof does not include the parts of the text (such
  as remarks, introductory passages, and auxiliary results in 
  Chapter~\ref{sec:further}) that are
  not strictly a part of the proof of the Kepler conjecture.
\item There are three specific sections of this book that describe
  the relationship between the text and the computer portions of
  the proof: Section~\ref{sec:hypermap-algorithm} (Hypermap Algorithm), 
  Section~\ref{sec:weight} (Main Estimate), and Section~\ref{sec:tsp} (Linear Programs).
  They have not yet been formalized.
\item There are still a few lemmas scattered throughout in the book that remain to be formalized.
These lemmas are listed at~\cite{website:FlyspeckProject}.
Combined, these scattered lemmas amount to fewer than 350 lines of English proof text.
\end{enumerate}
\item The formal verification of the hypermap generation program is complete~\cite{Nipkow:2005:Tame}.
\item The formal verification of linear programs is at a stage of 
quasi-completion~\cite{Obua:2005:Thesis} and~\cite{Solovyev:LP}.
 The technology to verify linear programs in HOL Light has been developed and 
  and efficient implementation has been made.
  There remains a the matter of connecting up the program and executing it.
\item The technology to make formal verifications of nonlinear inequalities over the
field of real numbers has been designed and implemented.   These are slow verifications,
and the software must be further optimized, before the verification of large collections
of inequalities becomes practical.  This research will appear in a forthcoming doctoral 
thesis by A. Solovyev
at the University of Pittsburgh.  Earlier work on formal proofs of nonlinear inequalities
is mentioned in~\cite{HHMNOZ}.
\end{enumerate}


The Flyspeck project has been a large team effort, and my contributions have
been a fraction of the whole.  The formal verifications have been made
over a period of years by a number of formal-proof experts.  This
appendix cites their contributions. The principal formalizer of each
chapter is indicated with an asterisk.

\def\x#1{\text{#1}}

\[
\begin{array}{lll}
\x{\bf Text~Chapter} & \x{\bf Author~of~formalization~work}
\vspace{4pt}\\
\x{Trigonometry} & \x{Nguyen Quang Truong\ast; Rute, Jason;}
\\&\x{Harrison, John; Vu Khac Ky; Trieu Thi Diep}\\
\x{Volume} & \x{Harrison, John\ast; Nguyen Tat Thang}\\
\x{Hypermap} &  \x{Tran Nam Trung}\ast\\
\x{Fan} & \x{Hoang Le Truong\ast; Harrison, John}\\
\x{Packing} & \x{Solovyev, Alexey\ast; Vu Khac Ky\ast;}\\
&\x{Nguyen Tat Thang; Hales, Thomas}\\
\x{Local Fan} & \x{Nguyen Quang Truong\ast}; \x{Hoang Le Truong\ast}\\
\x{Tame Hypermap} & \x{Solovyev, Alexey\ast; Dat Tat Dang;}\\
&\x{Diep Trieu Thi; Thanh Vu; Quyen Anh Vuong}\vspace{8pt}\\
\x{\bf Code-Verification} & \x{\bf Author~of~formalization~work}
\vspace{4pt}\\
\x{Hypermap Generation} & \x{Nipkow, Tobias\ast; Bauer, Gertrud\ast}\\
\x{Linear Programs} & \x{Obua, Stephen\ast; Solovyev, Alexey\ast}\\
\x{Nonlinear Inequalities} &  \x{Solovyev, Alexey\ast}\\
\end{array}
\]


\[
\begin{array}{lll}

\end{array}
\]


\newpage