%

\chapter{Cyclic Fan}\label{sec:cyclic}


\begin{summary}
At various times, it is useful to focus attention on a single face in a fan.  This leads to a construction of a what is called the localization of a fan along a face.  The localization is again a fan.  Special properties of the localized fan are captured in the notion of a cyclic fan.  The upper bound $2\pi$ on the perimeter of a cyclic fan is established.


To focus attention on a particular face $F$ of a fan $(V,E)$, it is useful to disregard all the vertices
and edges of the fan, except for the vertices and edges that constitute $F$.  This pared down fan
is an example of a {\it cyclic fan}, which is defined below.  It can be viewed as a ``localization'' of
the fan along a face $F$.  

A theory of deformations of cyclic fans is developed.  Sufficient conditions are stated for the deformation of a cyclic fan to remain a cyclic fan.
\end{summary}


\section{Localization}


\subsection{basics}


\begin{definition}[cyclic fan]  A triple $(V,E,F)$ is a {\it cyclic fan} if the following conditions hold.
\begin{nomerate} 
\item \case{fan} $(V,E)$ is a fan;
\item \case{face} $F$ is a face of $H = \op{hyp}(V,E)$;
\item \case{cyclic hypermap} $H$ is isomorphic to $H_{2k}$, where $k = \card(F)$;
\item \case{angle} $\op{azim}(x)\le \pi$ for all darts $x\in F$; and
\item \case{wedge} $V\subset \bWdart(x)$ for all $x\in F$.
%\item If $\{\v,\w\}\in E$, then $\{\orz,\v,\w\}$ is not collinear. %% part of def of fan.
\end{nomerate}
\end{definition}
\indy{Index}{fan!cyclic}%

The intersection of $X(V,E)$ with the unit sphere is a spherical polygon, when $(V,E,F)$ is a cyclic fan.  The spherical polygon gives a visual representation of the cyclic fan. The choice of $F$ distinguishes the ``interior'' of the polygon from its exterior.  The final two conditions are convexity constraints.  



\begin{lemma}\guid{WRGCVDR}\rating{ZZ}  
For any cyclic fan $(V,E,F)$, there is a bijection from $F$ onto $V$ given by
$$
 (\v,\w) \mapsto \v
$$
Moreover, write $\v\mapsto(\v,\rho\v)$ for the inverse map. 
Then $\rho:V\to V$ is a cyclic permutation.
%\begin{itemize}
%\item $(\v,\rho \v)\in D$, for all $\v\in V$; and
%\item 
%$$f(\v,\w) = (\w,\sigma(\w)^{-1}\v)= (\w,\rho \w)$$ for all $(\v,\w)\in F$.
%\end{itemize}
\end{lemma}
\indy{Index}{cyclic permutation}%

\begin{proof}  The map from the set of nodes to a face is a bijection for the cyclic hypermap $H_{2k}$, hence it is a bijection for cyclic fans.

For all $(\v,\rho \v)\in F$,
$$f(\v,\rho \v) = (\rho\v,\rho^2\v),$$
so that the order of $\rho$ on $V$ is the order of $f$ on $F$, which is $k=\card(F)$.  Thus, $\rho$ is a cyclic permutation of $V$ of order $k=\card(V)$.
\end{proof}

\begin{definition}[$\rho$,~$\v$]  For any cyclic fan $(V,E,F)$, 
write $\v:F\to V$ and $\rho:V\to V$ for the bijections of the preceding lemma.
\end{definition}
\indy{Notation}{ZZrho@$\rho$}%
\indy{Notation}{v@$\v$}%

\begin{definition}[interior angle,~$\angle$,~$\bWdart$]
For any cyclic fan $(V,E,F)$,
 write
$$
\angle(\v) = \op{azim}((\v,\rho\v)),
$$
for all $\v\in V$.  This is the {\it interior} angle of the cyclic fan at $\v$.
Also, write
$$
\Wdart(F,\v) = \Wdart((\v,\rho \v)),\quad \bWdart(F,\v) =\bWdart((\v,\rho \v)).
$$
\indy{Notation}{1@$\angle$}%
\indy{Notation}{Wdart@$\Wdart(F,\v)$}%
\indy{Notation}{Wdart@$\bWdart(F,\v)$}%
%\indy{Notation}{\v@$\v(x)$}%
\end{definition}






\begin{definition}[localization]  Let $(V,E)$ be a fan, and let $F$ be a face of $\op{hyp}(V,E)$.  
Let 
$$
\begin{array}{rll}
V' &= \{\v\in V \mid \exists~\w\in V.~(\v,\w)\in F\}.\\
E' &= \{\{\v,\w\} \in E\mid (\v,\w)\in F\}.
\end{array}
$$
The pair $(V',E')$ is called the {\it localization} of $(V,E)$ along $F$.
\end{definition}
\indy{Index}{localization}%


\begin{lemma}[localization]\guid{LVDUCXU}\rating{ZZ}\label{lemma:localization}  
Let $(V,E)$ be any fan, and let $F$ be a face of its hypermap.  
Then the localization $(V',E')$ is a fan, and $F$ is a face of $\op{hyp}(V',E')$.  Moreover, the hypermap $\op{hyp}(V',E')$ is isomorphic
to $H_{2k}$, where $k= \card(F)$.
 Finally, the values $\op{azim}(x)$ and $\Wdart(x)$  do not depend
whether they computed relative to $\op{hyp}(V,E)$ or to $\op{hyp}(V',E')$, for all $x\in F$.
%That is, 
%$$
%\begin{array}{lll}
%\op{azim}(V,E,x) &=\op{azim}(V',E',x)\\
%\Wdart(V,E,x) &=\Wdart(V',E',x)\\
%\end{array}
%$$
\end{lemma}



\begin{proof}
The proof that  $(V',E')$ is a fan is a sequence of simple verifications based on
the techniques of Remark~\ref{remark:fan-verify}.  The details are left to the reader.

The dart set $D'$ of $\op{hyp}(V',E')$ is naturally identified with the disjoint union $F\cup F'$, where
$F = \{(\v,\w) \in F\}$ and $F'=\{(\w,\v) \mid (\v,\w)\in F\}$.  Under this identification, $F$ is a face of $\op{hyp}(V',E')$.  The edge map $e(\v,\w)= (\w,\v)$ is a bijection of $F$ onto $F'$.   Pick any $x\in F$ and
define a bijection from the disjoint union of two copies of the cyclic group $Z_k$ onto $D'$  by 
the pair of maps
$$
i \mapsto f^i x,\quad\text{ and } i\mapsto e f^i x.
$$
This bijection extends to an isomorphism of hypermaps $H_{2k}$ onto $\op{hyp}(V',E')$.

The proof that the $\op{azim}(x)$ and $\Wdart(x)$ do not depend on the choice of fan is a consequence of their definitions:
$$
\begin{array}{lll}
\op{azim}(x) &= \op{azim}(\orz,\v,\w,\sigma(\v,\w)),\\
\Wdart(x) &= \Wdart(\orz,\v,\w,\sigma(\v,\w)).\\
\end{array}
$$
where $x = (\v,\w)$.    It is enough to check that $\sigma(\v,\w)\in E'(\v)$. 
But $\{\sigma(\v,\w),\v\}\in F$, so this is indeed the case.
%In particular $\op{azim}(x)\le\pi$ follow by the assumptions of the lemma.
\end{proof}



\subsection{geometric types}\label{sec:types}

\begin{definition}[generic,~lunar,~circular]
A cyclic fan $(V,E)$ is {\it generic} if for every $\{\v,\w\}\in E$
and every $\u\in V$, 
$$
C\{\v,\w\}\cap C^0_-\{\u\} = \emptyset.
$$
A cyclic fan is {\it circular} if there exists $\u\in V$ and $\{\v,\w\}\in E$ such that 
$$
C^0\{\v,\w\}\cap C^0_-\{\u\}\ne \emptyset.
$$
A cyclic fan is {\it lunar} with dependency $\v,\w$ if it is not circular and if $\v$ and $\w$ are distinct and dependent elements of $V$.
\end{definition}
\indy{Index}{generic}%
\indy{Index}{lunar}%
\indy{Index}{circular}%

\begin{lemma}\guid{CIZMRRH}\rating{ZZ} Every cyclic fan is
generic, lunar, or circular.  Moreover, these three properties are mutually exclusive.
\end{lemma}
\indy{Index}{fan!cyclic}%
\indy{Index}{cyclic fan}%
\indy{Index}{generic}%
\indy{Index}{lunar}%
\indy{Index}{circular}%

\begin{proof} If $(V,E,F)$ is not generic,  select some $\{\v,\w\}\in E$
and some $\u\in V$ such that
\begin{equation}\label{eqn:non-generic}
C\{\v,\w\}\cap C^0_-\{\u\} \ne \emptyset.
\end{equation}
Now $C\{\v,\w\} = C^0\{\v,\w\} \cup C\{\v\}\cup C\{\w\}$.  
If, for some such triple $(\u,\v,\w)$, the intersection~(\ref{eqn:non-generic}) meets $C^0\{\v,\w\}$, then the cyclic fan is circular.  
Otherwise, the cyclic fan is lunar. 
\end{proof}

\begin{lemma}\guid{LDURDPN}\rating{ZZ}  \label{lemma:coplanar}
Assume that $\u$ and $\w\in\ring{R}^3$ are independent, as well as $\u$ and $\v$.  Then $\op{azim}(\orz,\u,\v,\w)=\pi$
if and only if there exists a plane $A$ such that
$S=\{\orz,\u,\v,\w\}\subset A$ and such that the line $\op{aff}\{\orz,\u\}$ separates $\v$ from $\w$ in $A$.
\end{lemma}

\begin{proof}  The given azimuth angle is $\pi$ if and only if $\dih(\{\orz,\u\},\{\v,\w\})=\pi$.  This holds
exactly when $S$ is coplanar and the line $\op{aff}\{\orz,\u\}$ separates $\v$ from $\w$ in $A$.
\end{proof}

\begin{lemma}\guid{KOMWBWC}\rating{ZZ}\label{lemma:kom}
Let $(V,E,F)$ be a cyclic fan.  Let $k=\card(F)$.  Assume that for some $0<r\le k-1$ and some $\v\in V$, the set $U=\{\v,\rho \v,\ldots,\rho^r \v\}$ is contained in a plane $A$ passing through $\orz$.  Let $\e$ be the unit normal to $A$ in the direction $\v\times \rho \v$.  Then 
the set $U$ is cyclic with respect to $(\orz,\e)$, and the azimuth cycle $\sigma$ on $U$ is 
$$
\sigma \u = \begin{cases} \rho \u, & \u\ne \rho^r\v,\\ \v, & \u = \rho^r\v.\end{cases}
$$
\end{lemma}

\begin{proof} 
Write $\v_i = \rho^i \v$, for $i=0,\ldots,r$.

\claim{ 
$(\v_i \times \v_{i+1})\cdot \e > 0$, for all $i\le r-1$. }
Indeed, the base case of an induction  $(\v_0\times \v_1)\cdot \e > 0$ holds by assumption.    
Assume for a contradiction that the inequality holds for $i$, but not for $i+1$.
Then 
$$\op{aff}^0_+(\{\orz,\v_{i+1}\},\v_i) = \op{aff}^0_+(\{\orz,\v_{i+1}\},\v_{i+2}).$$ 
This forces $C^0\{\v_i,\v_{i+1}\}$ to meet $C^0\{\v_{i+1},\v_{i+2}\}$, which
is contrary to the definition of a fan.  Thus, the claim holds.

The fact that $U$ is cyclic follows trivially from the fact that $U$ is contained in a plane $A$ through $\orz$ and that $\e$ is orthogonal to $A$.

\claim{For all $0\le i \le r-1$,  $\sigma \v_i = \v_{i+1}$.}
Otherwise, there is some 
$$\u \in (U\setminus \{\v_i,\v_{i+1}\}) \cap W^0(\orz,\e,\v_i,\v_{i+1}) \cap A.
%  ~~\subset~~ 
$$
However, this intersection is a subset of $C^0\{\v_i,\v_{i+1}\}$,
and $\u\in C^0\{\v_i,\v_{i+1}\}$ is contrary to the property \case{intersection} of fans.  The result follows.
%The membership of $\u$ in the rightmost term is contrary to the definition of fan.  
%Let $\e_3$ be the unit vector in the direction $\v\times \u$.  Let $\e_1$ be the unit vector in the direction $\u$.
%Let $\e_2 = \e_3 \times \e_1$.  The coordinates of $\u,\v,\w$ with respect to the orthonormal frame $(\e_1,\e_2,\e_3)$ take the form
%$$
%\begin{array}{rllrl}
%\v &= * \e_1 + a \e_2,  &\quad & a &< 0\\
%\u &= a' \e_1, &\quad & a' &>0\\
%\w&= * \e_1 + a'' \e_2, &\quad & a'' &>0\\
%\end{array}
%$$
%From this representation it is clear that $\v\times \u$ points in the same direction as $\u\times \w$.
%The set $\{\v,\u,\w\}$ is clearly cyclic and the counterclockwise cycle $\sigma$ in the $\{\e_1,\e_2\}$ plane
%takes $\v$ to $\u$ and $\u$ to $\w$.
\end{proof}

\begin{lemma}\guid{OZQVSFF}\rating{ZZ} \label{lemma:A}  Let $(V,E,F)$ be a cyclic fan and let
$\u,\v,\w\in V$ such that
\begin{itemize}
\item $\{\orz,\u,\v,\w\}$ is contained in a plane $A$; \vspace{3pt}
\item $\u,\w\not\in\op{aff}\{\orz,\v\}$; and \vspace{3pt}
\item $\op{aff}^0_+(\{\orz,\v\},\u) \ne \op{aff}^0_+(\{\orz,\v\},\w)$.
\end{itemize}
Then $\angle(\v) = \pi$.  Moreover, $\rho \v,\rho^{-1} \v\in A$.
\end{lemma}

\begin{proof} Let $x = (\v,\rho\v)\in F$.  
Order $\u$ and $\w$ so that
$$
\op{azim}(\orz,\v,\rho\v,\u) \le \op{azim}(\orz,\v,\rho\v,\w).
$$
By the definition of cyclic fan, by the conditions $\u,\v\in \bWdart(x)$, 
and by  Lemma~\ref{lemma:coplanar},
%By the assumptions, $\dih(\{\orz,\v\},\{\u,\w\})=\pi$.  Since
%$\u,\w\in \bWdart(x)$, it follows that
%$$\pi = \dih(\{\orz,\v\},\{\u,\w\}) \le \op{azim}(x) = \angle(\v) \le \pi.$$
%The first conclusion follows.
$$
\begin{array}{rll}
0 &\le \op{azim}(\orz,\v,\rho\v,\u) \\
   &= \op{azim}(\orz,\v,\rho\v,\w) - \op{azim}(\orz,\v\,\u,\w)\\
    &= \op{azim}(\orz,\v,\rho\v,\w)-\pi \\
   &\le \op{azim}(\orz,\v,\rho\v,\rho^{-1}\v,\w) - \pi \\
   &=\op{azim}(x) - \pi \\
   &=\angle(\v)-\pi\\
   &\le 0. 
\end{array}
$$
Hence each inequality is equality.  In particular, $\angle(\v) = \pi$.
In particular, $0 =\op{azim}(\orz,\v,\rho\v,\u)$, so that 
$$\rho\v\in \op{aff}_+(\{\orz,\v\},\u) \subset A.$$
Similarly,
$$
\rho^{-1}\v\in \op{aff}_+(\{\orz,\v\},\w) \subset A.
$$
\end{proof}

If Lemma~\ref{lemma:A} can be applied once, then it can often be applied repeatedly along a chain.  For example, the conclusion of the lemma implies
that $\rho^{-1} \v \in A$.  In fact, by the definition of fan, 
$$\rho^{-1}\v \in A \setminus \op{aff}\{\orz,\v\} = \op{aff}^0_+(\{\orz,\v\},\u) \cup \op{aff}^0_+(\{\orz,\v\},\w).$$
Suppose that $\rho^{-1} \v$ lies in the first term of the union.  If $\u\ne \rho^{-1}\v$, then the assumptions of the lemma
are met for $\{\u,\rho^{-1} \v,\v\}$, giving the conclusion $\angle(\rho^{-1} \v)=\pi$ and $\rho^{-2}\v\in A$.  Repeating, we obtain a chain
$$
\pi=\angle(\v) = \angle(\rho^{-1} \v) = \cdots,
$$
with $\v,\rho^{-1}\v,\ldots\in A$.  Another chain can be constructed in the other direction $\v,\rho \v,\ldots$.
This process of chaining gives the following lemma.

\begin{lemma}[circular geometry]\guid{KCHMAMG}\rating{ZZ}\label{lemma:circular}
Let $(V,E,F)$ be a circular fan. Then
\begin{itemize}
\item $\angle(\v)=\pi$, for all $\v\in V$.
\item The set $V$ lies in a plane $A$ through $\orz$.
\item For some choice of unit vector $\e$ orthogonal to $A$, the set $V$ is cyclic with respect to
$(\orz,\e)$, and the azimuth cycle on $V$ coincides with $\rho:V\to V$.
Also,
$$
\op{azim}(\orz,\e,\v,\rho\v) = \op{arc}_V(\orz,\{\v,\rho \v\}).
$$
\end{itemize}
\end{lemma}

\begin{proof}  Let $\v, \u\in V$ be such that $C^0\{\u,\rho \u\}$ meets $C^0_-\{\v\}$.  Apply
Lemma~\ref{lemma:A} to $\{\u,\v,\rho \u\}$ to conclude that $\angle(\v)=\pi$ and some plane $A$ contains
$\{\orz,\u,\rho\u,\v,\rho \v,\rho^{-1} \v\}$.  If 
$$
\{\orz,\u,\rho \u,\v,\rho \v,\rho^{-1} \v\} \subset V\cap A,
$$
and $\w\in V\cap A$, then there exists $\w_1,\w_2\in (V\cap A)\setminus\{\w\}$ for which
the assumptions of Lemma~\ref{lemma:A} hold.  Then $\angle(\w)=\pi$ and
$\rho \w \in V\cap A$.  The set $V\cap A$ is therefore preserved by $\rho$.
By observing that $V$ is the only nonempty subset of $V$ that is preserved by $\rho$,
 it follows that $V\subset A$.  


By Lemma~\ref{lemma:coplanar}, $V$ is cyclic with respect to a unit vector $\e$ orthogonal to $A$.  The
azimuth cycle on $V$ is $\v \mapsto \rho \v$.
\end{proof}

\begin{lemma}[lunar geometry]\guid{HKIRPEP}\rating{ZZ}\label{lemma:lunar}
Let $(V,E,F)$ be a lunar fan with dependent vectors $\v,\w\in V$.  
Assume that $\rho^r \v = \w$, for
some $0< r < k$.
Then
\begin{itemize}
\item $\angle(\u) = \pi$, for all $\u\in V\setminus \{\v,\w\}$; \vspace{3pt}
\item $0< \angle(\v) = \angle(\w)\le \pi$; \vspace{3pt}
\item $V\cap \op{aff}_+(\{\orz,\v\},\rho \v) = \{\v,\rho \v,\ldots, \w\}$; \vspace{3pt}
\item $V \cap \op{aff}_+(\{\orz,\v\},\rho^{-1} \v) = \{\w,\rho \w,\ldots \v\}$;  \vspace{3pt}
\end{itemize}
\end{lemma}

\begin{proof}   Set $V_1 = \{\v,\rho \v,\ldots,\w\}$ and $V_2 = \{\w,\rho \w,\ldots,\v\}$. 
Let $\u\in V\setminus\{\v,\w\}$ be arbitrary.
Apply Lemma~\ref{lemma:A} to the set $\{\u,\v,\w\}$ to find that $\angle(\u)=\pi$ and
that $\{\orz,\u,\rho \u,\rho^{-1} \u\}$ belongs to a plane $A(\u)$.  Now $A(\u)$ and $A(\rho \u)$ are both the unique plane containing $\{\orz,\u,\rho \u\}$, hence $A(\u) = A(\rho \u)$ 
when $\rho \u\not\in \{\v,\w\}$.  By induction, there are planes $A_1, A_2$ such that $V_i\subset A_i$.  There is
an azimuth cycle $\sigma_i$ on $V_i$ such that $\sigma_i \u = \rho \u$, when $\u \in A_i\setminus \{\v,\w\}$.  

The angles $\angle(\v)$ and $\angle(\w)$ are both equal to the dihedral angle between these two half-planes.  In particular, $0<\angle(\v)=\angle(\w)\le\pi$.
\end{proof}




\begin{lemma}[monotonicity]\guid{EGHNAVX}\rating{ZZ}  \label{lemma:monotone}
Let $(V,E,F)$ be a cyclic fan. Fix $\v_0\in V$.  Assume that $\v_0$ and $\u$ are independent for all $\u\in V\setminus\{\v_0\}$.  For all $i$, set $\v_i = \rho^i \v_0$ and $\beta(i) = \op{azim}(\orz,\v_0,\v_1,\v_i)$.
Then
$$0=\beta(1)\le \beta(2)\le \cdots\le \beta(k-1)\le\pi.$$
Moreover, if $\beta(i)=0$ for some $1<i \le k-1$, then
$$
\angle(\v_1) = \cdots = \angle(\v_{i-1}) = \pi,
$$
and $\{\orz,\v_0,\ldots,\v_i\} \subset \op{aff}^0_+(\{\orz,\v_0\},\v_1)$.
Finally, if $\beta(i)=\beta(k-1)$ for some $1\le i<k-1$ then 
$$
\angle(\v_{i+1}) = \cdots = \angle(\v_{k-1}) = \pi,
$$
and $\{\orz,\v_i,\ldots,\v_{k-1}\} \subset \op{aff}^0_+(\{\orz,\v_0\},\v_{k-1})$.
\end{lemma}

\begin{proof}  
Pick an orthonormal frame and write the points $\v_j$ in spherical coordinates $(r_j,\theta_j,\phi_j)$.  In an appropriate orthonormal frame, $\phi_0=0$, and $\theta_j=\beta(j)$, for all $j$.  From $\v_j\in \bWdart(F,\v_0)$ and $\angle(\v_0)\le\pi$, it follows that $0\le\theta_j\le\theta_{k-1}\le\pi$ when $0\le j\le k-1$.

 One may assume by induction that $0\le \beta(1)\le\cdots\le \beta(i)$.  The condition
$$
\v_0\in \bWdart(F,\v_i)
$$
implies that
$$
0 \le \op{azim}(\orz,\v_i,\v_{i+1},\v_0)\le \op{azim}(\orz,\v_i,\v_{i+1},\v_{i-1})\le\pi.
$$
By Lemma~\ref{lemma:sim}, the resulting inequality
$$
\sin(\op{azim}(\orz,\v_i,\v_{i+1},\v_0))\ge 0
$$
reduces to a triple-product:
$$
(\v_0 \times \v_i)\cdot \v_{i+1}\ge 0.
$$
In spherical coordinates, this inequality becomes
$$
r_0r_ir_{i+1}\sin\phi_i\sin\phi_{i+1}\sin(\theta_{i+1}-\theta_i)\ge0.
$$
Under the independence assumption, $\sin\phi_i\ne0$ and $\sin\phi_{i+1}\ne0$ (when $0< i < k-1$).    These inequalities give $\theta_i\le\theta_{i+1}$ (with a small extra argument to exclude the degenerate case $\theta_{i+1}=0,\theta_i=\pi$).  The conclusion follows by induction.

Assume that $\beta(i)=\theta_i=0$, for some $i>1$.  
Then $\v_1,\ldots,\v_i$ all lie in the
half-plane $\op{aff}^0_+(\{\orz,\v_0\},\v_1)$.  In particular, they are coplanar.  A chaining argument based on Lemma~\ref{lemma:A} gives the result.

The final conclusion is similar.
\end{proof}



\subsection{deformation}\label{sec:deformation}

This section considers deformations of a cyclic fan $(V,E,F)$.

\begin{definition}[deformation]
A {\it deformation} of a cyclic fan $(V,E,F)$ over an interval $I\subset\ring{R}$
is a 
continuous function $\varphi:V\times I \to\ring{R}^3$ (in the discrete topology on $V$ and the product topology on $V\times I$).
%The deformation is {\it flat} if whenever
%$$\{\orz,\varphi(\rho^{-1}\v,t),\varphi(\v,t),\varphi(\rho \v,t)\}$$
%is coplanar for $t=a\in I$, then also the set is coplanar for all $t\in I$ such
%that $t\ge a$.
\end{definition}
\indy{Index}{deformation}%
\indy{Index}{fan!cyclic}%
%\indy{Index}{flat}%

%That is, a flat deformation is a deformation in which certain linear
%dependencies are ``sticky;''  once the dependency is formed, the dependency endures.

Write $\v(t)$ as an abbreviation of $\varphi(\v,t)$, for $t\in I$.  
Also, set
$$
\begin{array}{lll}
V(t)&=\{\v(t) \mid \v\in V\},\\
E(t)&=\{\{\v(t),\w(t)\}\mid \{\v,\w\}\in E\},\\
F(t)&= \{(\v(t),\w(t)) \mid  (\v,\w)\in F\}.
\end{array}
$$


A deformation does not require $(V(t),E(t),F(t))$ to be a cyclic fan for all $t\in I$, although this will often be the case. The permutation $\rho:V\to V$ gives $\varphi(\rho \v,t)\in V(t)$, for every $\v\in V$.  


\begin{example}[lunar deformation]\label{example:lunar}
Consider a lunar cyclic fan $(V,E,F)$ with dependent $\v,\w\in V$.   Pick an orthonormal frame and spherical coordinates such that $\phi(\v)=0$, $\phi(\w)=\pi$.  $\theta(\rho \v)=0$, and $\theta(\rho^{-1} \v)=\theta_{k-1}\le\pi$.  Consider the deformation $\varphi$ over 
$I = \{t \mid 0 \le t < 1\}$
such that radial $r$ and zenith $\phi$ coordinates of $\varphi(\u,t)$ are constant as functions of $t$,
and the azimuth angle $\theta$ of $\varphi(\u,t)$ equals

$$
\begin{cases} 
   (1-t) \theta_{k-1} & \text{if } \u\in \{\rho \w,\rho^2 \w,\ldots, \rho^{-1} \v\};\\
   0 & \text{if } \u\in \{\rho \v,\rho^2 \v,\ldots,\rho^{-1} \w\}.\\
\end{cases}
$$
%This deformation $(V(t),E(t),F(t))$ is a cyclic fan for all $t\in I$.
%The cardinality of $V$ is independent of $t\in I$.
Note that $(V(0),E(0),F(0)) = (V,E,F)$.
\end{example}
\indy{Index}{lunar}%
\indy{Index}{spherical coordinates}%

\begin{lemma}\guid{HZIYFIZ}\rating{ZZ}\label{lemma:lunar-deform} 
Let $(V,E,F)$ be a lunar cylic fan with dependency $\v,\w\in V$.  In the deformation described above, the triple $(V(t),E(t),F(t))$ is a lunar cyclic fan for all $t\in I$.  The cardinality of $V$ is independent of $t\in I$.
\end{lemma}

\begin{proof}  The proof consists of checking  the properties of a lunar cyclic fan, one by one.   The verification of the property \case{V} follows from the methods of Remark~\ref{remark:fan-verify}.

The map $\varphi(\cdot,t)$ is invertible, so that $V(t)$ is in bijection with $V$.  In particular, the cardinality does not depend on $t$.

\case{origin} The radial spherical coordinate is nonzero for every element of $V(t)$.  Hence $V(t)$ does not contain $\orz$.

\case{independence}  For $\{\v,\rho \v\}\in E$, the angle 
$\alpha=\op{arc}_V(\orz,\{\varphi(\v,t),\varphi(\rho \v,t)\})$
is independent of $t$.  The independence property is equivalent to $\alpha\ne0,\pi$.  Thus, the independence property for $E$ implies
the independence for $E(t)$.

\case{intersection} The points $V(t)$ are contained in the union of two half-planes $A_1,A_t$.  The deformation is the identity on $A_1$ and an isometry $A_0\to A_t$ on the second half-plane.  These bijections preserve the incidence relations of blades of the cones $C(\ee)$.  

\case{cyclic hypermap}~\case{face} The combinatorial properties of the hypermap do not depend on $t$.  In particular, the hypermap has face $F(t)$ and the hypermap is isomorphic to $H_{2k}$.

\case{angle} The azimuth angle $\angle(\varphi(\u,t))$ is fixed when $\u\ne \v,\w$ and is decreasing in $t$ when $\u\in \{\v,\w\}$.  Hence, the upper bound on the angle is preserved.

\case{wedge}  $V(t)\subset A_0\cup A_t$, where $A_0$ and $A_t$ are the
half-planes described above.  Now $\bWdart(F,\varphi(\u,t))$ is a half-space containing $A_0$ and $A_t$ when $\u\ne \v,\w$.  Also, $\bWdart(F,\varphi(\u,t))$ is a wedge with bounding half-planes $A_0$ and $A_t$ when $\u\in \{\v,\w\}$.  Hence
$V(t)\subset A_0\cup A_t\subset \bWdart(F,\varphi(\u,t))$ for all $\u\in V$.

\case{lunar} The points $\v,\w\in V(t)$ remain fixed and hence remain dependent.  Thus, the cyclic fan remains lunar.  
\end{proof}
\indy{Index}{angle!azimuth}%

Next  consider a deformation of a cyclic fan.  The following lemma gives a list of conditions that ensure that the deformation remains a cyclic fan throughout the deformation.


\begin{lemma}\guid{XRECQNS}\rating{ZZ}\label{lemma:fan-open}
Let $\varphi$ be a deformation of a  cyclic fan $(V,E,F)$ over an interval $I$.
Assume that $0\in I$ and that $\varphi(\v,0)=\v$ for all $\v\in V$.
For all $\v\in V$ assume that if $\{\orz,\rho^{-1}\v,\v,\rho\v\}$ is coplanar, then 
$$\{\orz,\varphi(\rho^{-1}\v,t),\varphi(\v,t),\varphi(\rho \v,t)\}$$
is coplanar for all sufficiently small $t$. 
If the property \case{wedge} is maintained for sufficiently small $t$, then
 exists $\epsilon>0$ such that $(V(t),E(t),F(t))$
is a  cyclic fan for all $t\in I\cap \leftopen-\epsilon,\epsilon\rightopen$.
Moreover, if $(V,E,F)$ is generic, then the property \case{wedge} is in fact maintained for sufficiently small $t$.  Finally, if $(V,E,F)$ is generic, then the deformation is also generic for sufficiently small $t$.
\end{lemma}

\begin{proof} Each of the properties of a generic cyclic fan will be examined in turn.

\case{V}:  The set $V(t)$ is the image of $V$ and is therefore finite and nonempty.  \case{origin} Since $\varphi$ is continuous and $\orz\not\in V$, it follows that $\orz\not\in V(t)$ for sufficiently small $t$.

\case{independence}: If $\v,\w$ are independent, then $\v(t)$ and $\w(t)$ are indepedent for sufficiently small $t$.

\case{intersection}: If $\ee \cap \ee'=\emptyset$, then $C(\ee)\cap S^2$ has a positive distance from $C(\ee')\cap S^2$.  Hence for sufficiently small times, the deformation of these sets remain disjoint.
If $\ee=\{\u,\v\}$ and $\ee'=\{\v,\w\}$ where $\u\ne\w$, then again the deformations of $C(\ee)\cap C(\ee')$ is $C(\{\v(t)\})$ for sufficiently small $t$.  The other cases are similar.

\case{face},~\case{cyclic hypermap}:  The azimuth cycle on $E(\v(t))$ is preserved; hence the combinatorial properties of the hypermap do not change when $t$ is sufficiently small.

\case{angle}: If $\op{azim}(x)<\pi$, then the inequality remains strict for sufficiently small $t$.  If $\op{azim}(x)=\pi$, then the coplanarity assumption of the lemma forces the equality to be preserved for all sufficiently small $t$.

\case{wedge}: The property $\u\in \Wdart(x)$ is an open condition.  It holds for sufficiently small $t$. Consider the case $\u\in \bWdart(x)\setminus \Wdart(x)$, where $x=  (\v_0,\rho\v_0)$.   Write $\v_i = \rho^i\v_0$ and pick $r\le k-1$ such that $\u = \v_r$.  The wedge property holds trivially when $r\in\{0,1,k-1\}$. Assume that $r\not\in\{0,1,k-1\}$. 
By the genericity assumption $\v_r$ and $\v_0$ are independent.  In the notation of Lemma~\ref{lemma:monotone},
$\beta(r) = 0$ or $\beta(r) = \beta(k-1)$.  This proof treats the case $\beta(r)=0$. (The case $\beta(r)=\beta(k-1)$ is similar.)   By Lemma~\ref{lemma:monotone}, 
$$0=\beta(1)=\beta(2)=\cdots=\beta(r),$$
and 
$$
\{\orz,\v_0,\v_1,\ldots,\v_r\} \subset \op{aff}^0_+(\{\orz,\v_0\},\v_1).
$$
In particular, the set is coplanar.   By the coplanarity assumption of the lemma,  $\{\orz,\v_0(t),\v_1(t),\ldots,\v_r(t)\}$ is coplanar for
sufficiently small $t$.  In fact, the condition $\v_r(t)\not\in \op{aff}_+(\{\orz,\v_0(t)\})$ is an open condition, so that $\v_r(t)\in \op{aff}^0_+(\{\orz,\v_0(t)\},\v_1(t))$ for sufficiently small $t$.  This half-plane is the bounding half-plane of $\bWdart(F,\v_0(t))$.  Hence $\u = \v_r\in \bWdart(F,\v_0(t))$ for sufficiently small $t$.

\case{generic}: Genericity is stated as open conditions $\v\not\in C\{\u,\w\}$.  These conditions continue to hold for sufficiently small $t$.
\end{proof}


%%%%%%%%%%%




\section{Perimeter}

%\subsection{perimeter}

\begin{definition}[perimeter]\label{lemma:perim}
Let $(V,E,F)$ be a cyclic fan.    Set
$$
\op{per}(V,E,F) = \sum_{i=0}^{k-1} \arc_V(\orz,\{\rho^i \v,\rho^{i+1} \v\}), 
$$
where  $k=\card(F)$.
This is easily seen to be independent of the choice of $\v\in V$.  Call $\op{per}$ the perimeter of the cyclic fan.
If $\v,\w\in V$ are distinct vertices, define the partial perimeter 
$$
\op{per}(V,E,F,\v,\w) = \sum_{i=0}^{r-1} \arc_V(\orz,\{\rho^i \v,\rho^{i+1} \v\}), 
$$
where $r$ is chosen so that $\w=\rho^r \v$ and $0<r\le k-1$.
\end{definition}
\indy{Index}{perimeter!cyclic fan}%
\indy{Index}{fan!cyclic}%
\indy{Notation}{per@$\op{per}$ (perimeter)}%



\begin{lemma}\guid{WSEWPCH}\tlabel{lemma:convex-hyp}\rating{400}
 The perimeter of every cyclic fan is at most $2\pi$.
\end{lemma}
\indy{Index}{fan!cyclic}%
\indy{Index}{perimeter}%

\begin{proof} 
\claim{If the cyclic fan is circular, then its  perimeter is $\op{per}(V,E,F) =2\pi$.}
Indeed, by Lemma~\ref{lemma:circular}, 
the arcs making up the perimeter all lie in a common plane.   
The azimuth cycle on $V$ coincides with $\rho:V\to V$.
The sum of the terms in the formula defining the perimeter is the sum of the azimuth angles in the azimuth cycle.  The sum is $2\pi$ by Lemma~\ref{lemma:2pi-sum}.


\claim{if the cyclic fan is lunar, then its  perimeter is $\op{per}(V,E,F) =2\pi$.}
Indeed, by Lemma~\ref{lemma:lunar}, the set $V$ is contained in the union of two half-planes.
The perimeter is the sum of arcs in a half-circle in the first half-plane plus
the sum of arcs in a half-circle in the second half-plane. This sum is $2\pi$.

Finally, assume that the cyclic fan is generic.
Suppose for a contradiction that the lemma is false.  Consider all counterexamples
that minimize the cardinality of $V$.  Among all such counterexamples, pick a counterexample with the smallest number of darts $x\in D$ such that $\op{azim}(x) = \pi$.

A cyclic fan $(V,E,F)$ is determined by $V$ and the cyclic permutation $\rho:V\to V$:  $E=\{\{\v,\rho \v\}\mid \v\in V\}$ and $F = \{(\v,\rho \v)\mid \v\in V\}$.

In this particular counterexample, if there is any dart $x=(\v,\w)\in F$ with $\op{azim}(x)=\pi$, then there is a new cyclic fan $(V',E',F')$ with $V' = V\setminus\{\v\}$ and $\rho':V'\to V'$ given by 
$$
\rho'(\u) = \begin{cases}
\rho(\u), & \text{if } \rho(\u)\ne \v;\\
\rho(\v), & \text{if }\rho(\u) = \v.\\
\end{cases}
$$
This is a cyclic fan with the same perimeter, contrary to the presumed minimality of the counterexample.  Thus $\op{azim}(x) <\pi$, for all $x\in F$.

If $\card(V) <3$, then the cyclic fan is circular or lunar, which has aleady been treated.  If $\card(V)=3$, then $V=\{\v_1,\v_2,\v_3\}$.  By the triangle inequality $\arc_V(\orz,\{\v_2,\v_3\}) \le \arc_V(\orz,\{\v_2,-\v_1\})+\arc_V(\orz,\{-\v_1,\v_3\})$.  Thus,
$$
\begin{array}{rll}
\op{per} &=\arc_V(\orz,\{\v_1,\v_2\}) + \arc_V(\orz,\{\v_2,\v_3\}) + \arc_V(\orz,\{\v_1,\v_3\})\\
&\le(\arc_V(\orz,\{\v_1,\v_2\})+\arc(\orz,\{\v_2,-\v_1\}))
   \\&\qquad\qquad+(\arc_V(\orz,\{\v_1,\v_3\})+\arc_V(\orz,\{-\v_1,\v_3\})) \\
   &= \pi+\pi.
\end{array}
$$

Now assume that $\card(V)\ge 4$.  Select $\v\in V$.  Consider a deformation of the cyclic fan $\varphi:V\times I \to \ring{R}^3$ that fixes $V\setminus\{\v\}$, and gives motion to $\v$:
$$
\varphi(\v,t) = \cos(t) \v - \sin(t) \rho \v.
$$
This is increasing in the perimeter.    For sufficiently small $t$,  it remains a generic cyclic fan (Lemma~\ref{lemma:fan-open}).  For sufficiently small $t$, the minimality gives $\angle(\varphi(\u,t))<\pi$.
Eventually, for some smallest $t=t_0$ the deformation is no longer a generic cyclic fan.
%  or for some $\u\in V$,  $\angle(\varphi(\u,t_0))=\pi$.  In the latter  case,  the minimality condition fails, and the result follows.  




\claim{The \case{independence} property holds at $t=t_0$}.  Indeed, it is enough to check independence when one of the points is $\varphi(\v,t_0)$ and the other is $\rho\v$ or $\rho^{-1}\v$.  The set $\{\orz,\rho\v,\v,\rho^i\v\}$ is not coplanar for $i=-1,-2$, because otherwise some interior angle is $\pi$, by Lemma~\ref{lemma:A}.   
Thus, the plane through $\{\orz,\v,\rho\v\}$ meets the plane through $\{\orz,\rho^{-2}\v,\rho^{-1}\v\}$ along a line $L$ through the origin.  
If the set $\{\orz,\rho^{-1}\v,\varphi(\v,t)\}$ is collinear, then that line is contained in $$\op{aff}\{\orz,\v,\rho\v\}\cap\op{aff}\{\orz,\rho^{-1}\v,\rho^{-2}\v\}=L,$$ which is impossible since $\rho^{-1}\v\not\in \op{aff}\{\orz,\v,\rho\v\}$.  Thus, $\rho^{-1}\v$ and $\varphi(\v,t)$ are independent.

\claim{The line $L$ meets the segment $\op{conv}^0\{\v,-\rho\v\}$.}  Otherwise,
it meets $\op{conv}\{\v,\rho\v\}$.  Then $\v$ and $\rho\v$ lie in distinct half-spaces
bounded by $\op{aff}\{\orz,\rho^{-1}\v,\rho^{-2}\v\}$.  Since $\bWdart(F,\rho^{-1}\v)$ is contained in one of the two half-spaces, the condition $\v,\rho\v\in \bWdart(F,\rho^{-1}\v)$ fails, and $(V,E,F)$ is not a cyclic fan.  This contradiction establishes the claim.

There is a time $t_1$ at which $\{\orz,\rho\v,\varphi(\v,t)\}$ is collinear, and a time $t_2$ at which $\varphi(\v,t)\in L$.  By the previous claim, $t_2<t_1$.
At time $t_2$,  some interior angle is $\pi$.  If $t_1\le t_0$, then $t_2 < t_0$, and this contradicts the observation that all interior angles of the fan are less than $\pi$.  So $t_1> t_0$. Thus, $\varphi(\v,t)$ and $\rho\v$ are independent. This gives independence.

\claim{The \case{intersection} property holds.}  If $\ee\subset V$, write $\ee(t) = \{\varphi(\u,t) \mid \u\in \ee\}$.   Let $W =W(\orz,\v,\rho\v,\rho^{-1}\v)$. The verification of the intersection property is based on the following facts (when $t>0$):
\begin{itemize} 
\item If $\v\not\in \ee$, then $C(\ee(t))=C(\ee)\subset W$.
\item $C^0\{\v(t)\} \cap W = \emptyset$.
\item $C^0\{\v(t),\rho^{-1}\v\}\cap W = \emptyset$.
\item $C\{\v(t),\rho\v\}\cap W = C\{\v,\rho\v\}.$
\item $C\{\v(t),\rho\v\}\cap C\{\v(t),\rho^{-1}\v\} = C\{\v(t)\}$.
\end{itemize}

\claim{At time $t=t_0$, the deformation is a cyclic fan.}
Otherwise, if the deformation fails to be a fan at time $t=t_0$, then the \case{independence} property,  \case{intersection} property, or the \case{cyclic} property fails.  
The first two properties have already been checked, so that the deformation must be a fan.  Furthermore, the conditions for a fan to be cyclic are closed conditions, so they must also hold. 

At $t=t_0$, the deformation is a cyclic fan that is not generic.   The perimeter bound follows  from previous cases.
\end{proof}

Here is a second proof of the same lemma.  It is conceptually much simpler, but more difficult to formalize.  It is based on polar polygons (a generalization of polar triangles to spherical polygons).

\begin{proof} A fan does not have any faces of cardinality less than three.
Every blade of the fan has radian measure less than $\pi$.  
\indy{Index}{polygon!polar}%

Consider the case of a spherical triangle.  If the edges of the
the triangle are $a_i$ and the angles of the polar
triangle are $\beta_i$, then $\beta_i=\pi-a_i$.
The the perimeter is 
$$a_1+a_2+a_3 = 2\pi - (\beta_1 -\beta_2 - \beta_3-\pi)= 2\pi-\op{sol} < 2\pi,$$
because the solid angle $\op{sol}$ of the polar triangle is always strictly positive.
\indy{Index}{triangle!spherical}%

Similarly, if the edges of the spherical polygon are
$a_i$, then the angles of the polar polygon are $\beta_i = \pi-a_i$.
The perimeter is
$$
a_1+\cdots+a_n  = 2\pi- \op{sol}< 2\pi,
$$
where $\op{sol} = 2\pi-\sum a_i$ is the solid angle of the polar polygon.
%~\cite[\p.261]{williamson:2008}.
\indy{Notation}{zzDelta $\op{sol}$ (area)}%
\end{proof}


\section{Special Fan}\label{sec:weight}  






\subsection{definition}

\begin{definition}[special~fan,~$\hm$]
Let $\hm = 1.26$.
A {\it special fan} is a tuple $(V,E,F,G)$, where
\begin{nomerate}
\item \case{packing} $V$ is a packing.  That is, for every $\v,\w\in V$, if $\norm{\v}{\w}<2$, then $\v=\w$.
\item \case{annulus} $V\subset \BB$.
\item \case{cyclic fan} $(V,E,F)$ is a cyclic fan.
\item \case{subset} $G\subset E$.
\item \case{g norm} If $\{\v,\w\}\in G$, then $\norm{\v}{\w}=2\hm$.
\item \case{e norm} If $\{\v,\w\}\in E$, then $\norm{\v}{\w}\le 2\hm$.
\item \case{diagonal} For all distinct elements $\v,\w\in V$, if
$\{\v,\w\}\not\in E$, then $$\norm{\v}{\w}\ge 2\hm.$$
\item \case{card} %$k=\card(F)$,
 Let      $s=\card(G)$ and $r=\card(E) - s = \card(F)-s$.  Then
$$0\le s \le 3,\quad\text{and}\quad3-s \le r \le 6 - 2s.$$
\end{nomerate}
The constants $r,s$ are called the {\it parameters} of the special fan.
\end{definition}
\indy{Index}{special fan}
\indy{Index}{parameters (of a special fan)}
\indy{Notation}{r (special fan parameter)}
\indy{Notation}{s (special fan parameter)}


\begin{definition}[d]
$$d(r,s) = \begin{cases}
    0.103 (2-s) + 0.2759 (r+2s-4), & r + 2s > 3\\
    0, & r + 2s \le 3.\\
    \end{cases}$$
\end{definition}
%d(3,0)=0, d(4,0)= 0.206; d(5,0)= 0.4819,   .7578
\indy{Notation}{d (lower bound for $\tau$)}

\begin{definition}[$\hm$,~$\tau$]
Let $(V,E,F)$ be a cyclic fan.  Set $\hm = 1.26$.  Set
$$
\tau(V,E,F) =\sum_{x\in F} \op{azim}(x)\left(1 + \dfrac{\sol_0}{\pi}  \dfrac{\normo{\v(x)}-2}{2\hm-2}\right) + \left(\pi+{\sol_0}\right) (2- k(F)),
$$
where  $\sol_0\approx0.551$ is the solid angle of a spherical equilateral triangle of side $\pi/3$, and $k(F)$ is the cardinality of $F$.
The function $\tau$ is defined on special fans by disregarding $G$:
$$
\tau(V,E,F,G) = \tau(V,E,F).
$$
\indy{Notation}{h0@$\hm$}
\indy{Notation}{zzt@$\tau$}
\end{definition}



\subsection{compactness}

Let 
$$
\begin{array}{lll}
V(\p) &= \{\p_{i}\mid i=0,\ldots,k-1\},\\
E(\p) &= \{\{\p_{i},\p_{i+1}\}\mid i=0,\ldots,k-1\},\\
F(\p) &= \{(\p_{i},\p_{i+1}) \mid i=0,\ldots,k-1\},\\
\end{array}
$$
where
 $\p:\{0,\ldots,k-1\}\to \ring{R}^3$ is any function, and
$\p_k = \p_0$.  (That is, for purposes of indexing, identify $\{0,\ldots,k-1\}$ with
the cyclic group $Z_k$.)
If $I\subset\{0,\ldots,k-1\}$ then set 
$$G(\p,I) = \{(\p_i ,\p_{i+1}) \mid i\in I\}$$

\begin{definition}[fan~datum]  Let $k\in\ring{N}$ and $I\subset \{0,\ldots,k-1\}$.
 A fan datum of shape $(k,I)$ is a function 
$\p:\{0,\ldots,k-1\}\to \BB$ such that
\begin{itemize}
\item \case{packing} For every $i,j$, if $\norm{\p_i}{\p_j}<2$, then $i=j$.
\item \case{i norm} If $i\in I$, then $\norm{\p_i}{\p_{i+1}}=2\hm$.
\item \case{e norm} For all $i$, $\norm{\p_i}{\p_{i+1}} \le 2\hm$.
\item \case{diagonal}  $\norm{\p_i}{\p_j} \ge 2\hm$ when $|i-j|>1$.
\item \case{card}  Let $s=\card(I)$ and $r=k-s$.  Then 
$$0\le s \le 3,\quad\text{and}\quad3-s \le r \le 6 - 2s.$$
\item \case{angle} $\op{azim}(\orz,\p_i,\p_{i+1},\p_{i-1})\le \pi$ for all $i$.
\item \case{wedge} $\p_j\in W(\orz,\p_i,\p_{i+1},\p_{i-1})$ for all $i,j$.
\end{itemize}
\end{definition}

\begin{lemma}\guid{VYNCGCO}\rating{ZZ}
If $\p$ is a fan datum of shape $(k,I)$, then
$$
(V(\p),E(\p),F(\p),G(\p,I))
$$
is a special fan.  Moreover, every special fan is equal to
$$
(V(\p),E(\p),F(\p),G(\p,I))
$$
for some fan datum $\p$ of some shape $(k,I)$.
\end{lemma}

\begin{proof}
\claim{Every special fan $(V,E,F,G)$ is equal to $(V(\p),E(\p),F(\p),G(\p,I))$
for some fan datum $\p$ of some shape $(k,I)$.}
Let $k=\card(F)$.  Fix any element $\u\in V$ and set $\p_i = \rho^i \u$.
Let $I = \{i\mid (\p_i,\p_{i+1}) \in G\}$.  Clearly $(V,E,F,G)=(V(\p),E(\p),F(\p),G(\p,I))$, and
$\p$ is a fan datum.

Now let $\p$ be any fan datum of shape $(k,I)$.

\claim{$(V(\p),E(\p))$
is a fan.}  The properties \case{V}, \case{origin}, and \case{independence} follow
by the methods of Remark~\ref{remark:fan-verify}.

\case{intersection}:  Some geometrical reasoning is required to establish the intersection property.   The case
$$
C\{\u\}\cap C\{\v\} = \{\orz\}
$$
follows from the strict triangle inequality 
$$
\normo{\u} \le 2\hm < 4 \le \normo{\v} + \norm{\u}{\v}.
$$
The other cases of the proof are based on the following 
two facts from Tarski arithmetic.
\begin{itemize}
\item Let $\{\v_0,\v_1,\v_2\}\subset \BB$ be a packing of three points.
Assume that $\norm{\v_1}{\v_2}\le 2\hm$.  Then
$C\{\v_0\}\cap C\{\v_1,\v_2\} = \{\orz\}$.
\item Let $\{\v_0,\v_1,\v_2,\v_3\}\subset \BB$ be a packing of four points.
Assume that $\norm{\v_1}{\v_3}\le 2\hm$ and $\norm{\v_2}{\v_4}\le 2\hm$.
Then $C\{\v_1,\v_3\}\cap C\{\v_2,\v_4\} = \{\orz\}$.
\end{itemize}

\claim{$(V(\p),E(\p),F(\p))$ is a cyclic fan.} The properties \case{wedge} and \case{angle} have been built into the definition of a fan datum.  The azmuth cycle $\sigma(\p_i)$ on $E(\p_i) = \{\p_{i+1},\p_{i-1}\}$ interchanges $\p_{i+1}$ with $\p_{i-1}$.  From this fact, the combinatorial properties \case{face} and \case{cyclic hypermap} are easily determined.  The bijection from the disjoint union of two copies of $Z_k$ onto the set of darts is given by the pair of maps
$$
i\mapsto (\p_i,\p_{i+1}),\quad i\mapsto (\p_{i+1},\p_i).
$$
This bijection extends to an isomorphism of hypermap $H_{2k}$ onto
$\op{hyp}(V(\p),E(\p))$.

\claim{$(V(\p),E(\p),F(\p),G(\p,I))$ is a special fan.} The properties of a special fan all follow trivially from the corresponding properties of a fan datum.
\end{proof}

The set of fan data of shape $(k,I)$ is a subspace of the topological space  $\BB^k$.


\begin{lemma}\guid{KFIIPLO}\rating{ZZ}
The space of fan data of shape $(k,I)$ is a compact metric space.  Moreover,
$$
\p \mapsto \tau(V(\p),E(\p),F(\p))
$$
is a continuous function on the space of fan data of shape $(k,I)$.
\end{lemma}

\begin{proof}  $\BB^k$ is a compact metric space and the constraints are all closed conditions.

The function $\tau$ is a polynomial in $\normo{\p_i}$ and $\op{azim}(\orz,\p_i,\p_{i+1},\p_{i-1})$.  The norm and azimuth angle are both continuous functions of $\p$.  
\end{proof}




\subsection{internal blades}


\begin{lemma}\guid{PGSQVBL}\rating{ZZ}  Let $(V,E,F)$ be a cyclic fan.   If $\v,\w\in V$ are independent, then $C\{\v,\w\} \subset \bWdart(x)$ for any dart $x\in F$.
\end{lemma}
\indy{Index}{fan!cyclic}%

\begin{proof}  This is an elementary consequence of the definitions, the cone shape of $\bWdart(x)$,  and the condition $V\subset \bWdart(x)$.
\end{proof}


\begin{lemma}[internal blades]\guid{YOLCBTG}\rating{ZZ}  \label{lemma:internal}
Let $(V,E,F)$ be a cyclic fan. 
Let $\v,\w\in V$ be independent.  Suppose that there exists $\v',\w'$ such
that $\angle(\v'),\angle(\w')<\pi$, where $\v,\v',\w,\w'$ are four distinct elements of
$V$ that appear in cyclic order.
Then $C^0\{\v,\w\}\subset \Wdart(x)$ for all $x\in F$.
%Pick a dart $x=(\v_0,\v_1)\in V$.  Set $\v_j = \rho^j \v_0$.  Assume that there are four darts $(y_1,y_2,y_3,y_4)$, $y_j = x_{j(j)}$, with
%$0\le j(1) < j(2) < j(3) < i(4)\le k-1$ 
%such that $\op{azim}(y_j) < \pi$, for $j=2,4$.  
%Then $C^0\{\v_{i(1)},\v_{i(3)}\} \subset \Wdart(x)$, for all $x\in F$.
\end{lemma}
\indy{Index}{blade!internal}%
\indy{Index}{internal blade}%
\indy{Index}{cyclic order}%

(To say that a sequence $\v_i$ of elements is in {\it cyclic order} means that
$\v_i = \rho^{j (i)}\v_0$, for some increasing function $j$ with range $\{0,\ldots,k-1\}$.)

\begin{proof}  Abbreviate $C^0 = C^0\{\v,\w\}$.  The first case to consider is $\v(x)=\v$.  For all $\p\in C^0\cap \bWdart(x)$, 
$$
0 \le \op{azim}(\orz,\v,\rho \v,\p) \le \op{azim}(\orz,\v,\rho \v,\rho^{-1} \v).
$$  
These inequalities are in fact strict.  If, for example $0 = \op{azim}(\orz,\v,\rho \v,\p)$, then
the set $\{\orz,\v,\rho \v,\w\}$ is coplanar.  Repeated application of Lemma~\ref{lemma:A} gives
$$
\angle(\v) = \angle(\rho \v) = \cdots = \angle(\rho^{-1} \w) = \pi,
$$
which is contrary to $\angle(\v') = \pi$.  
The strict inequalities imply $\p\in \Wdart(x)$ as desired.  The case $\v(x)=\w$ is similar.

Now assume that $\u=\v(x)\ne \v,\w$.  
By Lemma~\ref{lemma:A},  one may assume that $\{\orz,\u,\v,\w\}$ is not coplanar. 
(Otherwise, the contradiction $\angle(\v')=\pi$ or $\angle(\w')=\pi$ is reached.) Then
$$
\op{aff}\{\orz,\u,\v\}\cap C^0 \subset \op{aff}\{\orz,\u,\v\}\cap \op{aff}\{\orz,\v,\w\} \cap C^0 = \op{aff}\{\orz,\v\} \cap C^0 = \emptyset.
$$
Thus, $C^0$ is disjoint from $\op{aff}\{\orz,\u,\v\}$ and is similarly disjoint from $\op{aff}\{\orz,\u,\w\}$.  

We have the following facts:
$$
\v,\w\in W(\orz,\u,\v,\w),\quad C^0\{\v,\w\} \subset W(\orz,\u,\v,\w).
$$
Also,
$$
\begin{array}{rll}
C^0 &= C^0\cap W(\orz,\u,\v,\w) \\
     &\subset C^0 \cap (W^0(\orz,\u,\v,\w) \cup \op{aff}\{\orz,\u,\v\} \cup \op{aff}\{\orz,\u,\w\})\\
     &\subset C^0 \cap W^0(\orz,\u,\v,\w)\\
     &\subset \Wdart(x).
%\v,\w\in \barW &= \{\p \mid 0 \le \op{azim}(\orz,\u,\v,\p) \le \op{azim}(\orz,\u,\v,\w)\}.\\
%C^0 &\subset \barW\\
%\bar W &\subset W(\orz,\u,\v,\w) \cup \op{aff}\{\orz,\u,\v\} \cup \op{aff}\{\orz,\u,\w\}\\
%W(\orz,\u,\v,\w) & \subset \Wdart(x),
\end{array}
$$
%$$
%\Wdart(x) = \bWdart(x)\setminus (\op{aff}\{\orz,\u,\v\}\cup\op{aff}\{\orz,\u,\w\}).
%$$
\end{proof}

\begin{lemma}\guid{TECOXBM}\rating{ZZ}\label{lemma:2hm-slice}
Let $(V,E,F,G)$ be a special fan.  Let $\v,\w\in V$ be distinct elements such that $\norm{\v}{\w}=2\hm$ and such that $\{\v,\w\}\not\in E$.
Then $\v$ and $\w$ are independent, 
and $C^0\{\v,\w\}\subset \Wdart(x)$ for all $x\in F$.
\end{lemma}

\begin{proof} By Remark~\ref{remark:fan-verify}, independence follows from the strict triangle inequality
$$
\norm{\v}{\w} \le 2\hm < 2 + 2 \le \normo{\v} + \normo{\w}.
$$
Assume for a contradiction that the conclusion of the lemma is false.  Then by Lemma~\ref{lemma:internal}, all the intermediate internal angles between $\v$ and $\w$ are equal to $\pi$.  As a result, (after interchanging $\v$ and $\w$ if necessary), the set $\{\orz,\v,\rho\v,\rho^2\v,\ldots,\rho^r\v\}$ is planar, and
$$
\arc_V(\orz,\{\v,\rho^r\v\}) = \sum_{i=0}^{r-1} \arc_V(\orz,\{\rho^i\v,\rho^{i+1}\v\}),
$$
where $\rho^r \v = \w$ and $1 < r \le k-1$.
However, the left-hand side is at most
$$
\arc(2,2,2\hm) < 1.5,
$$
while the right-hand side is at least
$$
2\arc(2\hm,2\hm,2) > 1.5.
$$
This gives the desired contradiction.
\end{proof}

\subsection{slicing}

\begin{definition}[slice] Let $(V,E,F)$ be a cyclic fan.  Assume that $\v,\w\in V$ are
independent  and that
$(V,E')=(V,E\cup \{\{\v,\w\}\})$ is a fan.  Let $F'$ be the face of $\op{hyp}(V,E')$ 
containing the dart $(\w,\v)$.  Write
$$(V[\v,\w],E[\v,\w],F[\v,\w])$$
for the localization of $(V,E')$ along $F'$.  Explicitly,
$$
\begin{array}{lll}
V[\v,\w] &= \{\v,\rho \v,\rho^2 \v,\ldots,\w\};\\
E[\v,\w] &= \{\{\v,\rho \v\},\ldots,\{\rho^{-1}\w,\w\},\{\w,\v\}\};\\
F[\v,\w] &= \{(\v,\rho \v),(\rho \v,\rho^2 \v),\ldots,(\rho^{-1}\w,\w),(\w,\v)\}.
\end{array}
$$
The triple $(V[\v,\w],E[\v,\w],F[\v,\w])$ is called the {\it slice\/} of $(V,E,F)$ along
$(\v,\w)$.
\end{definition}
\indy{Index}{fan!cyclic}%
\indy{Index}{slice}
\indy{Notation}{1@$\cdot[\v,\w]$ (slicing a fan)}

To allow for more than one cyclic fan $(V,E,F)$,  expand the notation, writing $\angle(H,\v)$ for $\angle(\v)$ in the hypermap $H$.  Similarly, write $\Wdart(H,\v)$ for $\Wdart(x)$, and so forth.
\indy{Notation}{azimhv@$\op{azim}(H,\v)$}%
\indy{Notation}{wdart@$\Wdart$}%


\begin{lemma}[slicing]\guid{EJRCFJD}\rating{ZZ}\label{lemma:slice}  Let $(V,E,F)$ be a cyclic fan with hypermap $H$.  Pick $\v,\w\in V$. For each $\u\in \{\v,\w\}$, assume that $\u$ is independent of all elements of $V\setminus\{\u\}$.    Assume that $C^0\{\v,\w\}\subset \Wdart(x)$ for all darts $x\in F$.  Then
\begin{itemize}
\item $(V[\v,\w],E[\v,\w],F[\v,\w])$ and $(V[\w,\v],E[\w,\v],F[\w,\v])$ are cyclic fans.  
\item Let $H[\v,\w]$ and $H[\w,\v]$ be their hypermaps, respectively.  Let $g:V\to\ring{R}$ be any function.  Then
$$
\sum_{\v\in V} g(\v)\angle(H,\v) = \sum_{\v\in V[\v,\w]}g(\v)\angle(H[\v,\w],\v) + \sum_{\v\in V[\w,\v]}g(\v)\angle(H[\w,\v],\v).
$$
\end{itemize}
\end{lemma}
\indy{Index}{slice}%
\indy{Index}{fan!cyclic}%

\begin{proof} 
\claim{$(V,E')$ is a fan, where $E' = E\cup \{\{\v,\w\}\}$.}
 Indeed, except for the intersection property, all of the properties of a fan follow trivially from the fact that $(V,E)$ is a fan and
that $\v$ and $\w$ are independent.  (Note the similarity with Lemma~\ref{lemma:add-edge}.)
The intersection property also is trivial except in the case $\ee=\{\v,\w\}$ and $\ee'\setminus \ee\ne \emptyset$.  Pick $\u\in \ee'\setminus\ee$.  It follows from the node
partition of Lemma~\ref{lemma:disjoint} that
$$
\begin{array}{lll}
C(\ee) \cap C(\ee') &= (C(\v) \cap C(\ee')) \cup (C(\w)\cap C(\ee')) \\
 &= C(\{\v\}\cap \ee') \cup C(\{\w\}\cap \ee') \\
 &= C(\{\v,\w\}\cap \ee').
\end{array}
$$
The intersection property thus holds and $(V,E')$ is a fan.

It follows by Lemma~\ref{lemma:localization} that $(V[\v,\w],E[\v,\w])$ is a fan.

The second conclusion of the lemma follows from the following identities.
If $\u\ne \v,\w$ with $\u\in V[\v,\w]$, then $\u\not\in V[\w,\v]$ and 
\begin{equation}
\Wdart(H,\u)=\Wdart(H[\v,\w],\u),\quad \angle(H,\u) = \angle(H[\v,\w],\u).
\end{equation}
If $\u\in\{\v,\w\}$, then 
$\angle(H,\u)=\angle(H[\v,\w],\u) +\angle(H[\w,\v],\u)$.

Finally, it remains to be seen that the fan is cyclic.  Lemma~\ref{lemma:localization} already shows that the hypermap is isomorphic to $H_{2k}$. and that $F[\v,\w]$ can be identified with a face.  The condition $V[\v,\w]\subset \bWdart(x)$ follows from the fact that the angles $\beta(i)$ are increasing in Lemma~\ref{lemma:monotone}.
\end{proof}



The slicing procedure can also be applied to a special fan $(V,E,F,G)$.
Abbreviate the slices as
$$
\begin{array}{lll}
(V',E',F')&=(V[\v,\w],E[\v,\w],F[\v,\w]),\quad\text{and}\\
 (V'',E'',F'')&= (V[\w,\v],E[\w,\v],F[\w,\v]).
\end{array}
$$
Both edge sets $E'$ and $E''$ contain $\{\v,\w\}$.  The sets $G'$ and $G''$ can be defined
as 
$$
\begin{array}{lll}
G' &= \{\{\v,\w\}\} \cup (E'\cap G).\\
G'' &= \{\{\v,\w\}\} \cup (E''\cap G).\\
\end{array}
$$ 

Let $(r',s')$ and $(r'',s'')$ be the parameters for $(V',E',F',G')$ and $(V'',E'',F'',G'')$, respectively.
Set $k=r+s$, $k'=r'+s'$, and $k''=r''+s''$.

\begin{lemma}\guid{IXZYRSY}\rating{ZZ}\label{lemma:param-add}  
Let $(V,E,F,G)$ be a special fan (with parameters $s,r$).  
Pick distinct elements $\v,\w\in V$  such that $\{\v,\w\}\not\in E$.
Assume that $\norm{\v}{\w}=2\hm$.
For each $\u\in \{\v,\w\}$, assume that $\u$ is independent of all elements of $V\setminus\{\u\}$.    
%Assume that $C^0\{\v,\w\}\subset \Wdart(x)$ for all darts $x\in F$. 
Then $(V',E',F',G')$ and $(V',E',F',G')$ are special fans.  Moreover,
the parameters satisfy the relations
$$
k'+k'' = k + 2,\quad s'+s'' = s + 2,\quad r'+r''=r.
$$
Finally,
$$
\tau(V,E,F)= \tau(V'',E'',F'') +\tau(V',E',F').
$$
\end{lemma}

By interchanging $\v$ and $\w$, the lemma also asserts that $(V'',E'',F'',G'')$ is a special fan.

\begin{proof}  Each of the defining properties of a special fan will be considered in turn.  By Lemma~\ref{lemma:2hm-slice}, if $\{v,w\}\not\in E$, then $C^0\{v,w\}\subset \Wdart(x)$ for all darts $x\in F$.

The properties \case{packing}, \case{annulus}, \case{diagonal}, \case{subset}, \case{g norm}, and \case{e norm} follow directly from definitions and the corresonding properties for $(V,E,F,G)$.
The property \case{cyclic fan} follows from Lemma~\ref{lemma:slice}.

The relations between the constants $k,s,r$ for the various fans follows directly from the construction.
For example, there are two darts more in $F''\cup F'$ than in $F$.  

\claim{Property~\case{card} holds.}  Indeed, recall that each face in the hypermap of a fan has at least three darts.  Hence $3\le k$, $3\le k'$, and $3\le k''$.
The sets $G'$ and $G''$ contain $\{\v,\w\}$.  Hence $1\le s'$ and $1\le s''$.
From $0\le r\le 6 - 2s$, it follows that $s\le 3$.  If $s=3$, then $r=0$, and $k=r+s=3$.  This means that $F$ is a triangle.  For some ordering of the pair, $x=(\v,\w)$ is a dart in $F$.
Hence  $C^0\{\v,\w\} \not\subset \Wdart(x)$.  This is contrary to hypothesis.  Therefore, $s\le 2$.

Now for the verifications.
$$0\le s' = s + 2 - s'' \le s+1\le 3.$$
$$3-s'\le k'-s' = r'.$$
$$k' = k + 2 - k'' \le k-1.$$
$$
r'= k'+s' - 2 s' \le (k-1) + (s+1) - 2s' =k+s - 2s' = r + 2s -2s' \le 6 - 2s'.
$$
This completes the proof of property~\case{card}.

The additivity of $\tau$ follows directly from the azimuth angle estimates in Lemma~\ref{lemma:slice}.
\end{proof}


\section{Minimality}

%\subsection{minimality}


\begin{definition}[minimal~fan,~$k_{min}$,~$\tau^d_{min}$]
Let $k_{min}$ be the minimum of $r+s$ over
all special fans $(V,E,F,G)$ such that 
\begin{equation}\label{eqn:kmin}
\tau(V,E,F) < d (r,s),
\end{equation}
where $(r,s)$ are the parameters of the special fan.
If no special fan exists that satisfies the inequality, then set $k_{min}=0$.
Let $\tau^d_{min}$ be the infimum of $\tau(V,E,F)-d(r,s)$ over all special fans
$(V,E,F,G)$ whose parameters $(r,s)$ satisfy $r+s=k_{min}$.  If no special fans
exist with parameters $r+s=k_{min}$, then set $\tau^d_{min}=0$.
Any special fan $(V,E,F,G)$ with parameter $r+s=k_{min}$ such that
$$
\tau^d_{min}= \tau(V,E,F)-d(r,s)
$$
is a {\it minimal}  fan.
\end{definition}
\indy{Index}{fan!minimal}
\indy{Index}{minimal fan}


\begin{lemma}\guid{ADKOXQY}\rating{ZZ}\label{lemma:c-bound}
There is a constant $c>2\hm$ such that for every minimal fan $(V,E,F,G)$
and every distinct $\v,\w\in V$, either $\{\v,\w\}\in E$ or $\norm{\v}{\w}\ge c$.
\end{lemma}

\begin{proof} Pick a sequence of fan data $\p$ such that the associated
fan is a minimal fan and such that 
$$
\min_{|i-j|>1} \norm{\p_i}{\p_j}
$$
is tending to the minimal value $c$.  By passing to s subsequence, we may
assume without loss of generality that every term in the sequence has the same shape $(k_{min},I)$ for some $I$.  The sequence then lies in a compact metric space.  Passing again to a subsequence, we may assume without loss of generality that the sequence converges.  The limiting value is a fan datum whose associated minimal fan $(V,E,F,G)$ satisfies
$$
\norm{\v}{\w}=c,
$$
for some $\v,\w\in V$ such that $\{\v,\w\}\not\in E$.

Suppose for a contradiction that $c=2\hm$.  Then by Lemma~\ref{lemma:2hm-slice}, $C^0\{\v,\w\}\subset \Wdart(x)$, for all $x\in F$.
The fan can be sliced along the $\{\v,\w\}$.  
The parameters for the new pieces satisfy:
$$
k' = k+2 - k'' \le k-1.
$$
The function $d$ is additive over slices.  For some constants $d_1$ and $d_2$,
\begin{equation}\label{eqn:drs}
\begin{array}{lll}
 d(r,s) &= d_1 (2 - s) + d_2 (r + 2 s-4) \\
&= d_1 (2-s') + d_2 (r'+2 s'-4) + d_1 (2-s'') + d_2 (r''+2s''-4)\\
&= d(r',s') + d(r'',s''). \\
\end{array}
\end{equation}
By Lemma~\ref{lemma:param-add}, one of the two resulting fans satisfies inequality~\ref{eqn:kmin}.  This is contrary to the choice of $k_{min}$ in the definition of minimality.
\end{proof}





\subsection{genericity}

\begin{lemma}\guid{RRAJQBH}\rating{ZZ}\label{lemma:circular-nonmin}
Every minimal fan is generic.
\end{lemma}

\begin{proof}
By the classification of cyclic fans, it is enough to show that the fan is not circular and not lunar.

\claim{In a minimal fan, $\op{azim}(x)<\pi$ for some $x\in F$.}  Otherwise,
the result follows from the following estimates:
$$
\begin{array}{lll}
\tau(V,E,F) &=\sum_{x\in F} \op{azim}(x)\left(1 + \dfrac{\sol_0}{\pi}  \dfrac{\normo{v}-2}{2\hm-2}\right) + \left(\pi+{\sol_0}\right) (2- k(F))\\
 &\ge \sum_{x\in F} \pi + \left(\pi+{\sol_0}\right) (2- k(F))\\
 &= 2\pi + 2\sol_0 - k(F) \sol_0\\
 &\ge 2\pi - 4\sol_0\\
&> 0.7578\\
&=0.103 (2) + 0.2759 (2)\\
&\ge 0.103 (2-s) + 0.2759 (r+2s-4) \\ 
&= d(r,s)\\
\end{array}
$$
This proves the claim.


Section~\ref{sec:deformation} develops a theory of deformations of cyclic fans $(V,E,F)$.  Recall that a deformation is  continuous function $\varphi:V\times I\to\ring{R}^3$.  A deformation determines sets $V(t)$, $E(t)$, $F(t)$, and $G(t)$ for each $t\in I$.


\claim{A minimal fan is not lunar.}  Otherwise,
suppose for a contradiction, that $(V,E,F)$ is lunar, special, and
minimal.  Let $\v,\w\in V$ be dependent.   By the previous claim, we may assume without loss of generality that $\angle(\v)=\angle(\w)<\pi$.   Example~\ref{example:lunar} describes a deformation of the lunar fan $(V,E,F)$
over $I=\leftclosed0,1\rightopen$.  The deformation fixes $\normo{\u}$
and is non-increasing in $\angle(\u(t))$, for $\u\in V$.
From the defining formula for $\tau$, it follows that the function $\tau(V(t),E(t),F(t))$ is a decreasing function of $t$.

\claim{For sufficiently small positive $t$, the deformation is a special fan.}   The deformation is a cyclic fan for small positive $t$ by Lemma~\ref{lemma:fan-open}.   The property~\case{diagonal} of special fans holds by Lemma~\ref{lemma:c-bound}.  The other properties of a special fan follow immediately from the construction. 

However, the function $\tau$ attains its minimum at $(V,E,F)$ and hence
$\tau(V(t),E(t),F(t))$ cannot be decreasing in $t$.  This contradiction proves the claim that a minimal fan is not lunar.
\end{proof}

\subsection{irreducibility}

\begin{definition}[extremal,~minimal] Let $(V,E,F,G)$ be a special fan and let $\{\v,\w\}\in E$.
The edge $\{\v,\w\}\in E$ is {\it $G$-extremal} if $\{\v,\w\}\in G$ or
$$
\norm{\v}{\w}\in\{2,2\hm\}.
$$
The edge $\{\v,\w\}\in E$ is {\it $G$-minimal} if $\{\v,\w\}\in G$ or
$$
\norm{\v}{\w}=2.
$$
\end{definition}
\indy{Index}{extremal ($G$-extremal edge)}
\indy{Index}{minimal ($G$-minimal edge)}

\begin{definition} Let $(V,E,F)$ be a cyclic fan.
If $\angle(\v)=\pi$, then $\v$ is {\it flat}.
\end{definition}
\indy{Index}{flat (vertex)}

\begin{definition}[irreducible]  Let $(V,E,F,G)$ be a special fan and let $\v\in V$.  
Write $\v_i  = \rho^i \v$.
The special fan is {\it irreducible} at $\v$
if the following properties hold. 
\begin{itemize}
\item \case{extreme edge} For every $\w\in E(\v)$,  the edge $\{\v,\w\}\in E$ is $G$-extremal.
\item \case{flat exists}  If none of $\v_1,\v_2,\v_3,\v_4$ is flat, then $r+s\le 4$. 
% $r+s=4$. %, and $\norm{\u}{\w}\in\{ 2,2\hm}$ for all $\{\u,\w\}\in E$.
\item \case{no triple flat} At least one of $\v_0$, $\v_1$, and $\v_2$ is not flat.
\item \case{balance} If $\v$ is flat and if $\{\u,\v\},\{\w,\v\}\in E\setminus G$, then
$$
\norm{\u}{\v} = \norm{\w}{\v}.
$$
\item \case{g flat} If $\v_1$ and $\v_2$ are both flat, then 
$$G\cap \{\{\v_0,\v_1\},\{\v_1,\v_2\},\{\v_2,\v_3\}\} = \emptyset.$$
\item \case{flat middle} If $\v_1$ and $\v_2$ are both flat, then
$$
\norm{\v_1}{\v_2} = 2.
$$
\item \case{minimal vertex} Either $\normo{\v}=2$ or there exists $\w\in E(\v)$ such
that $\{\v,\w\}$ is $G$-minimal.
\item \case{minimal vertex flat} If $\v$ is flat, and $\{\u,\v\}\in E\setminus G$, then
$\norm{\u}{\v}=2$ or $\normo{\v}=2$.
\item \case{flat extremal} If $\v_1$ and $\v_2$ are both flat, then either $\v_1$ or $\v_2$
has extremal norm:
$$\normo{\v_1}\in \{2,2\hm\}\quad\text{ or }\quad \normo{\v_2}\in \{2,2\hm\}.$$
\item \case{extremal vertex}  If $\v_0$, $\v_1$, and $\v_2$ are not flat, then
$$
\normo{\v_1}\in \{2,2\hm\}.
$$
\item \case{flat extremal vertex}  If $\v_1$ is flat, but none of $\v_0,\v_2,\v_3$ is flat, then
$$
\normo{\v_1}\in \{2,2\hm\}\quad\text{ or }\quad\normo{\v_2}\in \{2,2\hm\}.
$$
\item \case{flat extremal vertex sym}  If $\v_2$ is flat, but none of $\v_0,\v_1,\v_3$ is flat, then
$$
\normo{\v_1}\in \{2,2\hm\}\quad\text{ or }\quad\normo{\v_2}\in \{2,2\hm\}.
$$
\end{itemize}
\end{definition}
\indy{Index}{irreducible}



\begin{lemma}\guid{KDKAGRS}\rating{ZZ}\label{lemma:min-irred}
Every minimal fan $(V,E,F,G)$ is irreducible at every $\v\in V$.
\end{lemma}

The proof appears in the next subsection as a series of small verifications.

\subsection{features}

 If there are $i$ consecutive flat vertices $\v_1,\ldots,\v_i$, then there are  $i+2$ corresponding vertices that lie in a plane $A$ through the origin: $\{\v_0,\ldots,\v_{i+1}\}\subset A$.


\begin{lemma}\guid{TESVAFW}
Every minimal fan $(V,E,F,G)$ satisfies property \case{no triple flat} of irreducibility at every $\v\in V$.  
\end{lemma}

\begin{proof}  Three flat vertices produces an arc
$$
\arc_V(0,\{\v_0,\v_4\}) = \op{per}(V,E,F,\v_0,\v_4) \ge4\arc(2\hm,2\hm,2) > \pi,
$$
but the remaining edges  have combined length at most
$$
\arc_V(0,\{\v_0,\v_4\})\le 2\arc(2,2,2\hm) < \pi.
$$
\end{proof}


\begin{lemma}\guid{SDCCMGA}\rating{ZZ}
Every minimal fan $(V,E,F,G)$ satisfies property \case{g flat} of irreducibility at every $\v\in V$.
\end{lemma}

\begin{proof}  Argue by contradiction.  From the constraints on $r$ and $s$, if $G\ne\emptyset$, then $s>0$ and $r+s\le 5$. 

If $r+s\le 4$, then $V = \{\v,\rho \v,\rho^2\v,\rho^3\v\}$.  A plane $A$ contains $V\cup\{\orz\}$.   This is not generic.

Thus, $r+s=5$.  A plane $A$ contains $\{\orz,\rho^{-1}\v,\v,\rho\v,\rho^2\v\}$. 
We obtain a contradiction by computing $\arc_V(\orz,\rho^{-1}\v,\rho^2\v)$ in two ways.  On the one hand, it is equal to the partial perimeter from $\rho^{-1}\v$ to $\rho^2 \v$.  One the other hand,
it is estimated by the triangle inequality, applied to the other two edges:
$$
\begin{array}{lll}
\arc_V(\rho^{-1}\v,\rho^2\v)
&\le\arc_V(\rho^2\v,\rho^3\v) + \arc_V(\rho^3\v,\rho^{-1}\v) \\
&\le\arc(2,2,2\hm)+\arc(y,2,2\hm)\\
&<\arc(y,2\hm,2)+\arc(2\hm,2\hm,2) +\arc(2\hm,2\hm,2\hm)\\
&\le\arc_V(\rho^{-1}\v,\rho^2\v).
\end{array}
$$
where $y=\normo{\rho^{-1} \v}$.
This is a contradiction.
\end{proof}

\begin{lemma}\guid{BAWWPPB}\rating{ZZ}
Every minimal fan $(V,E,F,G)$ satisfies property \case{extreme edge} of irreducibility for
every $\v\in V$.
\end{lemma}

\begin{proof} XX (Lexell, etc.)
\end{proof}


\begin{lemma}\guid{PZFSEVR}\rating{ZZ}
Every minimal fan $(V,E,F,G)$ satisfies property \case{minimal vertex} of irreducibility  for
every $\v\in V$.
\end{lemma}

\begin{proof} If the conclusion is false, then for sufficiently small positive $a$,  the deformation $\varphi$ over $[0,a]$ given by
$$
\varphi(\u,t) =
\begin{cases}
\u, & \text{if }\u \ne \v,\\
(1-t) \v, & \text{if }\u = \v\\
\end{cases}
$$
decreases $\tau$.  It is a deformation of special fans.  This shows that $(V,E,F,G)$ is not minimal.
\end{proof}

\begin{lemma}\guid{JLXMEJN}\rating{ZZ}   %case minimal vertex flat
Every minimal fan $(V,E,F,G)$ satisfies property \case{minimal vertex flat} of irreducibility  for
every $\v\in V$.
%\case{minimal vertex flat} If $v$ is flat, and $\{u,v\}\in E\setminus G$, then
%$\norm{\u}{\v}=2$ or $\normo{\v}=2$.
\end{lemma}

\begin{proof} 
Let $A$ be the plane containing $\{\orz,\rho^{-1}\v,\v,\rho\v\}$.  Assume that $\{\u,\v\}\in E\setminus G$.  Let $\u\ne \w\in E(\v)$.  Then $\u,\v,\w\in A$. 

If the conclusion is false, then for sufficiently small positive $a$, let $\varphi$ be the  deformation over $[0,a]$ that fixes all elements of $V$ other than $\v$ and moves $\v$ in the plane $A$ around the circle of radius $\norm{\w}{\v}$ with center $\w$.  The direction of the circular motion can be chosen to decrease $\tau$.  It is a deformation of special fans.  This shows that $(V,E,F,G)$ is not minimal.
\end{proof}

\begin{lemma}\guid{TPKKQOL}\rating{ZZ}
Every minimal fan $(V,E,F,G)$ satisfies property \case{balance}
of irreducibility at every $\v\in V$.
%For every edge $\{\v,\w\}\in E$ in a minimal fan $(V,E,F,G)$, 
%$$\norm{\v}{\w}\in\{2,2\hm\}.$$
\end{lemma}

\begin{proof}  Assume for a contradiction that the property fails.  In view of properties \case{extreme edge} and \case{minimal vertex flat}, after possibly swapping $\u$ and $\w$, we may assume without loss of generality that
$$
\normo{\v}=2,\quad \norm{\u}{\v}=2,\quad \norm{\w}{\v}=2\hm.
$$
For sufficiently small positive $a$, let $\varphi$ be the deformation over $[0,a]$
that fixes all elements of $V$ other than $\v$ and moves $\v$ in the plane of
$\{\orz,\u,\v,\w\}$ around a circle of radius $2$ with center at the origin.  The direction of the deformation can be chosen to be increasing in $\norm{\u}{\v}$.  The function $\tau$ is constant along the deformation.  The deformation carries minimal fans to minimal fans.  However, the deformed fan does not satisfy property \case{extreme edge}.  This is a contradiction.
\end{proof}




\begin{lemma}\guid{CFJSRQH}  Every minimal fan $(V,E,F,G)$ has property \case{flat middle} at every $\v\in V$.
\end{lemma}

\begin{proof} %Write $\v_i = \rho^{i-2} \v$.  
Assume for a contradiction tht the conclusion is false.  There is a plane $A$ that contains $\{\orz,\v_0,\v_1,\v_2,\v_3\}$.
By property \case{g flat}, none of the edges $\{\v_0,\v_1\}$, $\{\v_1,\v_2\}$, $\{\v_2,\v_3\}$ lies in $G$.  Furthermore, by \case{extreme edge} and \case{balance}, the distances $\norm{\v_i}{\v_{i+1}}$ are equal to one another and take values in $\{2,2\hm\}$, for $i=0,1,2$.  
By \case{minimal vertex}, if these distances are $2\hm$, then the partial perimeter from $\v_1$ to $\v_4$ is at least
\begin{equation}\label{eqn:3side}
\arc(y_1,2,2\hm)+\arc(2,2,2\hm)+\arc(2,y_4,2\hm),
\end{equation}
where $\normo{\v_i}=y_i$.
However, there are at most two other vertices ($\v_4$ and $\v_5$) and three other edges in the special fan.  By the triangle inequality, the sum of these three lengths is at most \eqn{eqn:3side}.
Thus, equality is obtained in the triangle inequality.  This implies that $\v_4$ and $\v_5$ lie in the plane $A$.  Thus, $V\subset A$.   This is not generic.
\end{proof}

\begin{lemma}\guid{OUCPLRI} 
Every minimal fan $(V,E,F,G)$ satisfies property \case{flat extremal} of irreducibility at every $\v\in V$.
\end{lemma}

\begin{proof}
Assume there are two adjacent flat angles $\v_1,\v_2$, so that there is a plane $A$ that contains $\{\orz,\v_0,\v_1,\v_2,\v_3\}$.
By previously established properties of irreducibility,
\indy{Index}{angle!flat}%
$\norm{\v_i}{\v_{i+1}}=2$, for $i=0,1,2$.
Let $y_i = \normo{\v_i}$.
%

\claim{There exists a deformation that fixes all $\u\ne \v_1,v_2$, that keeps $y_1+y_2$, $\norm{\v_0}{\v_1}$, $\norm{\v_2}{\v_3}$,  and $\tau$ constant, while increasing $\norm{\v_1}{\v_2}$.}  Indeed,
In fact, the partial perimeter $c$ is given by a sum of three terms:
  $$
  c=\sum_{i=0}^2\arc(y_i,y_{i+1},2).
  $$
Set
$$
g(t_1,t_2) = \arc(y_0,y_1+t_1,2) + \arc(y_1+t_1,y_2-t_1,2+t_2) + \arc(y_2-t_1,y_3,2) - c.
$$
Then $g(0,0)=0$.  

Use subscript notation for partial derivatives of $g$.  If $g_1(0,0) \ne 0$, then by the implicit function theorem there is a function $h(t_2)$ locally near $0$ such that $h(0)=0$ and $g(h(t_2),t_2)=0$.  Define a deformation that fixes $\u\ne\v_1,\v_2$ and moves $\v_1,\v_2$ in the fixed plane $\op{aff}\{\orz,\v_1,\v_2\}$ subject to the constraints $\norm{\v_1(t)}{\v_2(t)}=2+t$, $\normo{\v_1(t)} = 2+h(t)$, and $\normo{\v_1(t)} = 2-h(t)$.  The deformation is a special fan for sufficiently small $t$.  It keeps $\tau$ fixed.  Hence the deformation is also a minimal fan.  However, the existence of this deformation contradicts property \case{flat middle}.

On the other hand, if $ g_1(0,0) =0$, then a calculation gives
$$g_2(0,0) = \dfrac{2}{y_1y_2\sqrt{\ldots}} > 0.$$
Again by the implicit function theoreom there is a function $h$ locally near $0$ such that $h(0)=0$ and $g(t_1,h(t_1))=0$.   Define a deformation that fixes $\u\ne \v_1,\v_2$ and moves $\v_1$ and $\v_2$ within the fixed plane $\op{aff}\{\orz,\v_1,\v_2\}$ by the conditions $\normo{\v_1(t)} = 2 + t$, $\normo{\v_2(t)} = 2 - t$, and $\norm{\v_1(t)}{\v_2(t)}=2+h(t)$.  

By implicit differentiation,  $h'(0) = 0$ and 
$h''(0) = -g_{11}(0,0)/g_2(0,0)$.  Thus, $h''(0)$ and $g_{11}(0,0)$ have opposite signs.
The function $g(t_1,0)$ equals, up to a constant, the sum of three terms of the form
$\arc(s_1(t_1),s_2(t_1),2)$.
A calculation shows that the second derivative of each term with respect to $t_1$ is negative at $t_1=0$.
%% XX check this.
Thus, $h''(0)>0$ and for sufficiently small nonzero $t$, 
$\norm{\v_1(t)}{\v_2(t)}>2$.
The deformation is a special fan for sufficiently small $t$.  It keeps $\tau$ fixed.  Hence the deformation is also a minimal fan.  However, the existence of this deformation contradicts property \case{flat middle}.
\end{proof}







\begin{lemma}\guid{DWXPIHA}\rating{ZZ}
Every minimal fan $(V,E,F,G)$ satisfies property \case{extremal vertex} for every $\v\in V$.
\end{lemma}

\begin{proof} %\guid{DFSLRHA} 
Assume for a contradiction that the conclusion is false.
Consider the deformation $\varphi:V\times I\to V$ given by
$$
\varphi(\u,t) =
\begin{cases}
\v(t), & \text{if } \u=\v\\
\u, & \text{otherwise},\\
\end{cases}
$$
where $\v(t)$ is the unique point in $\op{aff}_+(\{0,\rho^{-1}\v,\rho \v\},\v)$ determined by the conditions:
$$
\norm{\v(t)}{\u} = \norm{\v}{\u}, \text{~when~} \{\u,\v\}\in E,\quad
\text{~and~}
\normo{\v(t)} = (1-t)\normo{\v}.
$$
For sufficiently small $t$, the deformation  is a special fan.
The function
$t\mapsto\tau(V(t),E(t),F(t))$ has no local minimum on $(-\epsilon,\epsilon)$ when $\epsilon$ is positive and sufficiently small.
(The proof is an interval arithmetic calculation over a four-dimensional space~\cite[cc:d2a]{hales:2009:nonlinear}.  %%cc:d2a
The calculation verifies  that the second derivative is negative whenever the derivative is zero.)  This is contrary to the minimality of $(V,E,F,G)$.
\end{proof}



%The same can be done when there is one flat vertex forming a linear series of length $2$, with compressed edges:

\begin{lemma} \guid{DCEETTF}\rating{ZZ}
Every minimal fan $(V,E,F,G)$ satisfies properties \case{flat extremal vertex} and property \case{flat extremal vertex sym} for every $\v\in V$.
\end{lemma}

\begin{proof}
The property \case{flat extremal vertex sym} is obtained from \case{flat extremal vertex} by symmetry $v_1\leftrightarrow v_2$, so it is enough to prove \case{flat extremal vertex}.  Assume for a contradiction that property \case{flat extremal vertex} fails.
Let $(V,E,F,G)$ be a minimal fan that fails.

Define a deformation $\varphi$ of the minimal fan that fixes $\u\ne \v_1,\v_2$ and that moves $\v_1,\v_2$ subject to the following constraints.
\begin{itemize}
\item $\{\orz,\v_0,\v_1(t),\v_2(t)\}$ is coplanar.
\item $\norm{\v_2(t)}{\v_3}$ is constant.
\item $\norm{\v_0}{\v_1(t)}$, $\norm{\v_1(t)}{\v_2(t)}$, and $\norm{\v_0(t)}{\v_2(t)}$ are constant.
\item $\normo{\v_2(t)} = \normo{\v_2} + t$.
\end{itemize}
These constraints uniquely determine the deformation. The deformation is a special fan for sufficiently small $t$.

\claim{The function
 $\tau(V(t),E(t),F(t))$  is twice continuously differentiable.  The function does not have a local minimum at $t=0$.}
Indeed, this follows from an interval arithmetic calculation over a five-dimensional space~\cite[cc:d2b]{hales:2009:nonlinear}. %%cc:d2b
The calculation shows that the function  has a negative second derivative at $t=0$ if the derivative is zero.

By the claim,  $\tau$ has no local minimum along the deformation.  This contradicts the assumption that $(V,E,F,G)$ is minimal.
\end{proof}



\begin{lemma}\guid{CMBZAOZ}\rating{ZZ}
Every minimal fan $(V,E,F,G)$ satisfies property \case{flat exists} of irreducibility for every $\v\in V$.
That is, if none of $\v,\rho \v,\rho^2 \v,\rho^3 \v\in V$ is flat, then $r+s\le 4$.
%then $r+s=4$. %, and $\norm{\v}{\w}= 2$, for all $\{\v,\w\}\in E$.
\end{lemma}

\begin{proof} 
Assume for a contradicton that the property fails for $(V,E,F,G)$.
In particular, the parameters satisfy $r+s>4$.
There are four consecutive vertices $\v_1,\v_2,\v_3,\v_4$ that are not flat.   By property \case{extreme edge},   each edge $\norm{\v_i}{\v_{i+1}}$ is $2$ or $2\hm$. Set $y_i = \normo{\v_i}$ and $y_{ij} = \norm{\v_i}{\v_j}$.   By \case{extreme vertex}, $y_i$ is $2$ or $2\hm$, for $i=2,3$.   The condition $r+s>4$ implies that $\{\v_1,\v_4\}\not\in E$ and $y_{14}\ge 2\hm$.

Define a deformation of $(V,E,F,G)$ by fixing $\u\ne \v_2,\v_3$ and moving $\v_2,\v_3$ according to the following constraints:
\begin{itemize}
\item $\norm{\v_1}{\v_2(t)}$, $\norm{\v_2(t)}{\v_3(t)}$, $\norm{\v_3(t)}{\v_4}$, and $\norm{\v_1}{\v_2(t)}$ are constant.
\item $\normo{\v_2(t)}$ and $\normo{\v_3(t)}$ are constant.
\item $\norm{\v_1}{\v_3(t)} = y_{13} + t$.
\end{itemize}
These constraints uniquely determine the deformation. The deformation is a special fan for sufficiently small $t$.

In this context, the function $\tau$ as a function of $t$ (with parameters $y_{13},y_1,y_4,y_{14}$) does not have a local minimum at $t=0$.
%% XX check this.
(This is an interval arithmetic calculation~\cite[cc:qua]{hales:2009:nonlinear} in four variables.) %% cc:quad tau-estimate, deform a quadrilateral.
This is contrary to the assumed minimality of $(V,E,F,G)$.
\end{proof}

%As a consequence, the minimum occurs when a new flat vertex forms.   Then eliminate any case with four consecutive vertices that are not flat (after treating the quadrilateral with four edges $y_{ij}=2$ as a special case).

\subsection{emptiness}

\begin{lemma}\guid{LPQUDGF}\rating{ZZ}\label{lemma:min-empty}  
The set of minimal fans is empty.
\end{lemma}

\begin{proof}  
Let 
$$X= \{(V,E,F,G,\v_0) \mid  (V,E,F,G) \text{ is minimal and } \v_0\in V\}.
$$  The strategy is to partition $X$ into smaller sets and show that each set is empty.

For each $(V,E,F,G,\v_0)\in X$, there is a unique fan datum $\v$ and shape $(k_{min},I)$ such
that $\v_i = \rho^i\v_0$ and $(V,E,F,G) = (V(\v),E(\v),F(\v),G(\v,I))$.  The index $i$ ranges over the cyclic group $Z_k$ where $k=k_{min}$.  

For each $(V,E,F,G,\v_0)$ with fan datum $\v$, define a function
$$
\op{vlabel}:Z_k \to \{\op{low},\op{mid},\op{high}\} \times \{\op{flat},\op{nonflat}\}
$$
(where $\op{low}$, $\op{mid}$, $\op{high}$, $\op{flat}$, and $\op{nonflat}$ are symbolic labels for elements in sets of cardinality three and two, respectively).
Define
$$
\op{vlabel}(i)_1 = \begin{cases}
   \op{low}, &\text{if } \normo{\v_i} = 2,\\
   \op{high}, &\text{if } \normo{\v_i} = 2\hm,\\
   \op{mid}, &\text{otherwise}
\end{cases}
$$
and
$$
\op{vlabel}(i)_2 = \begin{cases}
   \op{flat}, &\text{if $\v_i$ is flat},\\
   \op{nonflat}, &\text{otherwise}.
\end{cases}
$$

For each $(V,E,F,G,\v_0)$ with fan datum $\v$, define a function
$$
\op{elabel}:Z_k \to \{\op{low},\op{mid},\op{high}\} \times \{\op{g},\op{nong}\}
$$
(where $\op{low}$, $\op{mid}$, $\op{high}$, $\op{g}$, and $\op{nong}$ are symbolic labels for elements in sets of cardinality three and two, respectively).
Define
$$
\op{elabel}(i)_2 = \begin{cases}
   \op{low}, &\text{if } \norm{\v_i}{\v_{i+1}} = 2,\\
   \op{high}, &\text{if } \norm{\v_i}{\v_{i+1}} = 2\hm,\\
   \op{mid}, &\text{otherwise}
\end{cases}
$$
and
$$
\op{elabel}(i)_2 = \begin{cases}
   \op{g}, &\text{if $i\in I$},\\
   \op{nong}, &\text{otherwise}.
\end{cases}
$$

For given functions $\op{vlabel}$ and $\op{elabel}$, 
let $X(\op{vlabel},\op{elabel})$ be the subset of $X$ with the given functions.
There are finitely many labeling functions $\op{vlabel}$ and $\op{elabel}$.
Thus, we may enumerate all cases.

By Lemma~\ref{lemma:min-irred}, every minimal fan is irreducible at every point in $V$.
The labels have been designed in such a way that every fan in a given set
$X(\op{vlabel},\op{elabel})$ is irreducible or no fan in the set is irreducible.
Each set that does not contain an irreducible fan is empty.  We may filter the
set of labeling functions, discarding those that do not contain an irreducible fan.

If $f:Z_k\to Y$ is any function, then define the shift $f[j]:Z_k\to Y$ by
$f[j](i) = f(i+j)$.  It is clear that $X(\op{vlabel},\op{elabel})$ is empty
if and only if $X(\op{vlabel}[j],\op{elabel}[j])$, because $(V,E,F,G,\v_0)$ lies
in the first set if and only if $(V,E,F,G,\rho^j\v_0)$ lies in the second set.  Thus,
we may reduce the list of labeling functions so that they are listed up to a shift.

A short computer calculation shows that every irreducible set of labels is equal
to one of the following up to a shift:
\begin{itemize}
\item XX
\end{itemize}

The elements of a given set $X(\op{vlabel},\op{elabel})$ can be parametrized
by the lengths 
$$\normo{\v_i} \mid \op{vlabel}(i)_1 = \op{mid},$$
$$\norm{\v_i}{\v_{i+1}} \mid \op{elabel}(i)_1 = \op{mid},$$
and $k_{min}-3$ diagonals.  In every case, the number of such parameters is
at most six.    Interval arithmetic calculations %XX check
show in each case that
$$
\tau(V,E,F) \ge d(r,s),
$$
for every $(V,E,F,G,\v_0)\in X(\op{vlabel},\op{elabel})$.  (A separate calculation
has been made for each of the labeling functions in the list above.)
This inequality shows that none of the elements $(V,E,F,G,\v_0)$ is minimal.
This completes the proof that $X$ is empty.
 ``Praise to the emptiness that blanks out existence.'' -- Rumi % The essential Rumi page 21.
\end{proof}

\begin{corollary}\guid{JEJTVGB}\rating{ZZ}\label{lemma:empty-d}
$$
\tau(V,E,F) \ge d (r,s)
$$
for every special fan $(V,E,F,G)$, where $(r,s)$ are the parameters of $(V,E,F,G)$.
\end{corollary}

\begin{proof} 
The function $\p\mapsto \tau(V(\p),E(\p),F(\p),G(\p,I))$ is continuous on the space of fan data of shape $(k_{min},I)$

Assume for a contradiction that the conclusion is false.
Then $k_{min}>0$, because it is the cardinality of some nonempty set $V$.
There exists a sequence of fan data $\p_i$ of varying shapes $(k_{min},I_i)$ such that 
$\tau(V(\p_i),E(\p_i),F(\p_i))-d(r,s)$ tends to $\tau^d_{min}$.  By passing to a subsequence, we may assume without loss of generality that $I_i = I$ is independent of $i$.
By passing again to a subsequence in the compact 
metric space of fan data of shape $(k_{min},I)$, the sequence $\p_i$ converges to some fan datum $(\p,I)$.
It follows that $\tau^d_{min} = \tau(V,E,F)-d(r,s)$, where $(V,E,F,G)=(V(\p),E(\p),F(\p),G(\p,I))$.   
This is a minimal fan.
However, the set of minimal fans is empty.
\end{proof}


\section{Old Proof}


\begin{proof}
More than one of these can be combined in a single polygon.   Represent the sizes of the linear series in a polygon with $k$ sides as a partition of $k$.
The number of parts of the partition gives the number of sides of the polygon when each linear series is considered a single side. 
\indy{Index}{linear!series}%

Let us review the combinatorial possibilities.  In each case,  check that the inequality holds, to complete the proof of the main estimate.  Label each 
case by the partition.  If the partition is $(\mu_1,\mu_2,\ldots)$, 
index the vertices in the same order as the partition, with $\v_0$
the vertex occurring just before the first flat vertex.  For example,
if the partition is $(3,1,1)$, the polygon is a pentagon, flattened into
an effective triangle, with vertices $\v_0,\ldots,\v_4$, where $\v_1$ and $\v_2$
are the flat vertices.
\indy{Notation}{ZZmu@$\mu$ (partition)}%

The following constraints hold on the partition: it is
a partition of $k$, where $3\le k\le 6$; there
are at least three parts; the largest part is at most $3$;  there are
at most two consecutive parts that equal $1$, unless it is a special triangular case $(1,1,1)$ or
quadrilateral case $(1,1,1,1)$, discussed below.  In only one case
does the order of the parts matter: $(2,1,2,1)$ versus $(2,2,1,1)$.




\begin{itemize}
\item {\bf (1,1,1)}  This is a triangle.  There are three degrees of
freedom $y_i$.  Since this is arising from a polygon that was not originally a triangle,  assume that for at least one of the edges: $y_{ij}=2\hm$.
\item {\bf (1,1,1,1)}  This is a convex quadrilateral with four sides $y_{ij}=2$ and four extremal heights $y_i$.  There is one degree of freedom given by the diagonal.
\item {\bf (2,1,1)} This is a quadrilateral, flattened into an effective triangle.  There are two degrees of freedom: one of the three variables $y_0,y_1,y_2$ is extremal.  All of the other variables are extremal.
\item {\bf (3,1,1)}  This is a pentagon, flattened into an effective triangle.  There are three degrees of freedom $y_0,y_{03},y_3$ along the flattened side, and no freedom in the rest of the figure.  It follows that $y_{01}=y_{12}=y_{23}=2$.
\item {\bf (2,2,1)} This is a pentagon, flattened into an effective triangle.  The edge lengths $y_{i,i+1}$ are all extremal.  Because of height adjustments, there are three degrees of freedom in the heights $y_i$.
\item {\bf (3,2,1)} This is a hexagon, flattened into an effective triangle.  There are four degrees of freedom.
\item {\bf (2,2,2)}  This is a hexagon, flattened into an effective triangle.  There are six degrees of freedom, given by all heights $y_i$.
\item {\bf (2,1,2,1)} This is a hexagon, flattened into an effective quadrilateral.  There are five degrees of freedom: a diagonal $y_{03}$ to the quadrilateral and four independent heights.
\item {\bf (2,2,1,1)}  This is a hexagon, flattened into an effective quadrilateral.  There are four degrees of freedom: a diagonal $y_{03}$ to the quadrilateral and three independent heights.
\indy{Index}{triangle}%
\indy{Index}{quadrilateral}%
\indy{Index}{pentagon}%
\indy{Index}{hexagon}%
\end{itemize}


Interval arithmetic calculations~\cite[cc:tau]{hales:2009:nonlinear} %% cc:par partition cases for tau[r,s]. 
for each of these cases completes the proof.  The set of minimal fans is empty. 
\end{proof}
