% Author Thomas C. Hales
% Copyright Thomas C. Hales
% Format latex.
% 

%!TEX TS-program = latex    
%% This line is for TexShop. 

% Revision history. See svn.
% Document created Dec 6, 2002
% Revision started Jan 2007, from published DCG.

\documentclass[cup9a]{cupbook}
%% Cambridge University Press Macros from
%% https://authornet.cambridge.org/information/productionguide/laTex_files/

%

\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{latexsym}
\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{amsmath}
%\usepackage{mathidx}
\usepackage{makeidx}
\usepackage{multicol}
\usepackage{crop}
\usepackage{txfonts}
%\usepackage{pdfsync}  %for TexShop sync.
\usepackage[letterpaper,colorlinks=true,
            ps2pdf,hyperindex=true]{hyperref}
%\usepackage{mparhack} %http://www.tex.ac.uk/cgi-bin/texfaq2html?label=marginparside
\usepackage{multind}
\usepackage{url}






\input{flag}
\input{kepmacros}

 \crop
%\makeindex

% CUP BOOK CLASS SPECIFIC
\newtheorem{lemma}{Lemma}[chapter]
\newtheorem{definition}{Definition}[chapter]
\newtheorem{remark}{Remark}[chapter]
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{corollary}{Corollary}[chapter]
\newtheorem{example}{Example}[chapter]
\newtheorem{claim}{Claim}[chapter]
\newtheorem{notation}{Notation}[chapter]
\newtheorem{assumption}{Assumption}[chapter]
\newtheorem{interpretation}{Interpretation}[chapter]
\newtheorem{conjecture}{Conjecture}[chapter]
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\raggedbottom  % for now.
%\raggedright  % don't worry for now.

    \title{Flyspeck :
      %\\ \phantom{Hello}
      \\  A Blueprint for the Formal Proof of the Kepler Conjecture
      \\ (Marchal Version)}
    \author{Thomas C. Hales}
    \maketitle
    \frontmatter
    \tableofcontents
    \thanks

\mainmatter

\input{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%  \label{part:intro}
   \newpage

   \setcounter{chapter}{-1}
   \chapter{Preface}
   \label{XX}  % for missing references. XX.
   %\input{preface} %proofed
   
   

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  
    
    \input{trig}  
    \input{volume}
    %\input{hypermap}
    
    \chapter{Hypermap}
    Use the electrical dual charge proof%
    \footnote{\url{http://www.ics.uci.edu/~eppstein/junkyard/euler/all.html}} of the Euler formula for convex polyhedra (a combinatorial Morse function).  This does not seem to need an explicit invocation of the Jordan curve theorem, since the faces of the polyhedron are convex.
    
    
      
    The local index of a vertex is $2$ - number of min/max edges + number of min/max faces. The local index at the absolute max and min = 2 each.  The local index at any other vertex is $0$ (two of the local face elements are not a max or min).  Sum of local indices $=4$.
    
    The sum of the vertex contributions is $2V$, edge contributions $-2E$, and face contributions $2F$ (by convexity each face has a unique max and min). 
	Thus, the graph is planar: $2 = V-E+F$.

	Without loss of generality, the spherical net of edges formed by quasi-regular tetrahedra is 2-connected.  
    
    Conjecture: We can show that the quasi-regular tetrahedra unit sphere arcs belong to the set of edges of the convex hull of the projection of the points to the unit sphere.  
    
    Removing an edge that does not disconnect the graph does not change the Euler characteristic, so we can work our way from the planar graph formed by the convex hull to the planar graph formed by the quasi-regular tetrahedra.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \input{marchal}
    %\input{tame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    %\input{lin_prog} 
 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %\input{XX_issues}
    \input{backmatter}
    %%


\end{document}
