% Template for an AMS article style
\documentclass{article} %{llncs}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{alltt}
\usepackage{makeidx}
\usepackage{txfonts}
\usepackage[rflt]{floatflt}


\makeindex

% Math notation.
%\def\op#1{{\operatorname{#1}}}
%\newcommand{\ring}[1]{\mathbb{#1}}

% Flags
\def\verbose{f}
\def\leftalign{\text{\hbox to 8in{\hss}}}

\input{kepmacros}

\newtheorem{rx}{}[section]
\newtheorem{lemma}{Lemma}[rx]
\newtheorem{definition}{Definition}[rx]
\newtheorem{remark}{Remark}[rx]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[rx]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\raggedbottom

\title{The Dodecahedral Conjecture}
\author{Thomas C. Hales\ and Sean McLaughlin}
%\address{University of Pittsburgh, Carnegie Mellon University}
%\email{hales@pitt.edu, seanmcl@cmu.edu}
\maketitle

\begin{abstract}  
This article gives a proof of Fejes T\'oth's Dodechaedral Conjecture: The volume of a Voronoi polyhedron in a packing of balls of unit radius is at least the volume of a regular dodecahedron of unit inradius.
\end{abstract}
%\frontmatter
%\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%

%\mainmatter
\input{dodec}

\thanks{Research supported by NSF grant 0503447.   Many of the nonlinear inequalities for the proof
of the Dodecahedral conjecture were proved by computer by S. Ferguson.  We
wish to thank him for his generous support of this project.}

%%%%%

\begin{thebibliography}{HeHa94}

\bibitem[Bau06]{Bauer} G. Bauer, Formalizing Plane Graph
Theory -- Towards a Formalized Proof of the Kepler
Conjecture, dissertation, Technische
Universit\"t M\"unchen, 2006.

\bibitem[BN06]{BN06} G. Bauer, T. Nipkow, P. Schultz,
Flyspeck I: Tame Graphs, 2006.

\bibitem[Bez97]{Bezdek} K. Bezdek, {\it Isoperimetric inequalities and the
dodecahedral conjecture}, Int. J. Math., \emph{8/6} (1997), 759-780.

\bibitem[Bez04]{Bez04} K. Bezdek, Sphere packings in $3$-space,
Proc. of the COE Workshop on Sphere Packings, Kyushu Univ. Press 2004, 32--49.

\bibitem[Col76]{Col} G. E. Collins, Quantifier elimination for real closed fields by cylindrical 
algebraic decomposition. In Brakhage, H. (ed.), Second GI Conference on 
Automata Theory and Formal Languages, Volume 33 of Lecture Notes in Computer Science, Kaiserslautern, pp. 134Ð183. Springer-Verlag. 1976.

\bibitem[FT43]{Toth1} L. Fejes T\'oth, {\it \"Uber die dichteste
Kugellagerung}, Math Z., \emph{48} (1943), 676-684.

\bibitem[FT72]{Fej72} L. Fejes T\'oth, {\it Lagerungen in der Ebene auf der
    Kugel und im Raum}, second edition,
    Springer-Verlag, Berlin New York, 1972.

\bibitem[FT64]{Toth2} L. Fejes T\'oth, {\it Regular Figures}, pp.263-300,
Pergamon Press, Oxford-London-New York, 1964.


\bibitem[FH06]{DCG} S. Ferguson, T. Hales, A Proof of the
Kepler Conjecture, 
Discrete and Computational Geometry, 35:1, 2006.

\bibitem[Gol]{Float} D. Goldberg, What Every Computer Scientist Should
Know About Floating-Point Arithmetic, Computing Surveys, ACM, 1991.
http://docs.sun.com/source/806-3568/ncg\_goldberg.html.

\bibitem[Gon05]{Gon} G. Gonthier, A computer-checked proof of the Four
Colour Theorem, preprint 2005.

\bibitem[Har68]{Approx} John F. Hart et. al, Computer Approximations, John
Wiley and Sons, 1968.

\bibitem[HM02]{arx} T. Hales, S. McLaughlin,
A Proof of the Dodecahedral Conjecture (2002 version),  	arXiv:math/9811079v2 [math.MG].

\bibitem[Ha92]{Hales1} Thomas C. Hales, The sphere packing problem, Journal
of Computational and Applied Math, 44 (1992) 41--76.

\bibitem[Ha97]{Part1} Hales, {\it Sphere Packings 1}, Dis. and Comp. Geom., 
{\bf 17} (1997), 1-51.

\bibitem[Hal94]{Hal94} T. C. Hales, The status of the Kepler conjecture,
    {\it Math. Intelligencer} 16, no. 3, (1994), 47--58.

\bibitem[Ha03]{alg} Hales, {\it Some algorithms arising in the proof of the
	Kepler conjecture}, 
	Discrete and computational
geometry, 489--507, Algorithms Combin., 25, Springer, Berlin,
2003.
	
	\bibitem[Hal07]{Fl}   T. Hales, The Flyspeck Fact Sheet, 2003; revised
2007. http://code.google.com/p/flyspeck/wiki/FlyspeckFactSheet.
	
	\bibitem[Hal05c]{fly}  Thomas C. Hales, The Flyspeck Project, 
{\it Mathematics, Algorithms, Proofs}, Dagstuhl seminar proceedings, 
ed. T. Coquand, H. Lombardi, and M.-F. Roy,
IBFI 2005.
	





%\bibitem{HalesWeb} Thomas C. Hales, Packings,\hfill%\break\quad {\tt
%http://www.math.lsa.umich.edu/\~\relax hales/}

\bibitem[Hal05]{HalesWeb} Thomas C. Hales, Computer resources for the Kepler conjecture, \hfill\break
    http://www.math.princeton.edu/\~\!annals/KeplerConjecture/ (2005 snapshot).
    http://code.google.com/p/flyspeck/ (regularly maintained).
    %\hfill{\it http://www.math.pitt.edu/\~%
    %\relax thales/kepler98.html} \hfil\break
    % (The source code, inequalities,
    %and other computer data relating to the solution are also found
    %at {\it http://xxx.lanl.gov/abs/math/9811078v1}.)
    %% {Hal98A} Needs to be updated to this.
    
\bibitem[Har96]{Har} E. Harshbarger, Saturated Packings and Voronoi cells, http://www.ericharshbarger.org/voronoi.html, 1996.

\bibitem[Hor83]{Hor}  L. H\"ormander,
The Analysis of Linear Partial Differential
Operators II, Volume 257 of Grundlehren der
mathematischen Wissenschaften. Springer-Verlag,
1983.

\bibitem[Hs93]{Hsiang} W.-Y. Hsiang, {\it On the sphere packing problem and
the proof of Kepler's conjecture}, Int. J. Math., \emph{4/5} (1993),
739-831.

\bibitem[AH83]{Interval} G\"otz Alefeld and J\"urgen Herzeberger,
Introduction to Interval Computations, Academic Press, 1983.


\bibitem[754]{IEEE} IEEE Standard for Binary Floating-Point Arithmetic,
ANSI/IEEE Std. 754--1985.

\bibitem[Kea96]{Kea96} R. B. Kearfott, Rigorous Global Search: Continuous
Problems, Kluwer 1996.

\bibitem[Law97]{cfsqp} C. Lawrence, J. L. Zhou, A. L. Tits,  User's Guide for CFSQP Version 2.5:
   A C Code for Solving (Large Scale) Constrained Nonlinear (Minimax) Optimization Problems, Generating
   Iterates Satisfying All Inequality Constraints, 
   http://www.aemdesign.com/download-cfsqp/cfsqp-manual.pdf, 
   Institute for Systems Research, University of Maryland, Technical Report TR-94-16r1, 1997.


\bibitem[Le56]{Leech} Leech, John.  {\it The Problem of the Thirteen
Spheres}, Math. Gazette, Feb. 1956, 22-23.

\bibitem[McL08]{McLp} S. McLaughlin,
Objective Caml bindings for the CFSQP nonlinear optimization library, http://code.google.com/p/ocaml-cfsqp/, 2008.

\bibitem[Code]{code} S. McLaughlin,
(XX Sean: Pick a name and set up a permanent link)
Computer Resources for the Dodecahedral Conjecture,
http://www.XX\_Give\_URL, 2008.


\bibitem[Mu88]{Muder1} D.J. Muder, {\it Putting the best face on a Voronoi
polyhedron}, Proc. London Math. Soc., \emph{3/56} (1988), 329-348.




\bibitem[Mu93]{Muder2} D.J. Muder, {\it A new bound on the local density of
sphere packings}, Discrete and Comp. Geom., \emph{10} (1993), 351-375.

\bibitem[NB06]{BN}  T. Nipkow, G. Bauer, P. Schultz, Flyspeck I: Tame Graphs, in
{\it Lecture Notes in Computer Science}, 21--35, 4130 (2006). 

\bibitem[Ob08]{Ob}  S. Obua, Flyspeck II: The Basic Linear Programs,
thesis, TU Munich 2008.

\bibitem[Pr92]{Numerics} William H. Press et al., Numerical Recipes in C,
Chapter 20. Less-Numerical Algorithms, Second Edition, Cambridge
University Press, 1992.

\bibitem[Ro58]{Rogers} C.A. Rogers, {\it The packing of equal spheres},
J. London Math. Soc., \emph{3/8} (1958), 609-620.



\bibitem[Zu06]{Zu}  R. Zumkeller,  Formal Global Optimisation with
 Taylor Models, {\it Automated Reasoning -- IJCAR 2006},
 ed. U. Furbach and N. Shankar, LNCS, 408--422, 4130/2006.


\end{thebibliography}





\end{document}




