

\chapter{Chapter 11. Upright Quarters and Valued Objects}

Def: We defined cyclic set above.  We also have a {\it permuted set}, which
is a set $X$ together with a permutation $s$ of it.

We have seen that the boundary of a standard region region can be traversed
so that the region lies on the left.

Each edge of the standard region is the projection of an edge of
$U(D)$. Consider pairs $(v,w)$ where the segment from $v$ to $w$
is an edge of $U(D)$.  The indexing of the centered packing $D$
gives a pair $(i_v,i_w)$ that determines $(v,w)$.  Let $E$ be the
set of edge indices determined in this manner.  $E$ is a {\it
permutated set\/} with the action $s$ that sends $(i_v,i_w)$ to
$(i_w,i_u)$ where $(w,u)$ is the edge the follows that edge
$(v,w)$ in the left traversal of edges around $R$, with $R$ the
standard region to the left of the projection of $(v,w)$.  Since
$D$ determines $E$ and $s$, we may consider them part of the data
of a centered packing.

Def:  A standard cluster is a pair $(D,R)$ where $D$ is a centered
packing and $R$ is a standard region of $R$.

Def:  Let $\text{size}(D,R)$ be the number of sides of the
standard region counted with multiplicity.  The cluster is
exceptional if $\text{size}(D,R) > 4$.  The {\it corners} of
$(D,R)$ is the intersection $U(D) \cap C(R)$; that is, the set of
vertices of the centered packing lying in the cone over $R$.

Def: The {\it local } $V$-cell is
$$V_{D,R} = VC(D,0) \cap C(R).$$
The truncated local $V$-cell is
$$V_{D,R} = VC(D,0) \cap C(R) \cap B(0,t).$$

Often, we take $t = t_0$.

Def: (Valued object)  A valued object is a pair $(f,X)$ consisting
of a function $f:X \mapsto \ring{R}$.  The
{\it value\/} of $x\in X$ is $f x$.  If $Z$ is a finite
set of valued objects, an element of $Z$ is a function $a:Z \to -$,
where $X_z(a z)$ for all $Z z$.  The {\it total value} of $a$ is
$$\sum_Z (f_z (a z)).$$

Def: A {\it singleton valued object} is a value object of the form
$(f,\{x\})$.

Def: Let $D$ be a centered packing.  Let $S$ be an indexing set.
Let $\text{delete} D S$ be the centered packing with the vertices
$v_i$, $i \in S$ indexed by $S$ deleted, together with all the all
the indexing data that meets $S$.

L: Let $S$ be any indexing set, then $\text{delete} D S$ is again
a centered packing. (XX This needs to be checked.)

We create a large number of valued objects.  Some of the main estimates
of this paper come through the relations between these various valued
objects.

\begin{enumerate}
\item (GFQ) $\gamma$ on a subset of flat quarters \item (SFQ)
$\sigma$ on a flat quarter \item etc. other flat quarter, upright
quarters, qrtets. \item etc. on octahedra. \item $\sigma$ on a
centered packing \item $\sigma$ on a standard cluster $(D,R)$.
\item etc. Types A, B, C, C'. \item little delta at upright $v$.
\item scoring on standard region, etc. \item deformable
standard region, etc. \item LP bounds on sets, etc. \item
(ABSOLUTE BOUND) $I$ on $\ring{R}$.
\end{enumerate}

We consider all valued objects as an enumerated type in the sense of CAML
(check terminology).

We create a domination tree.  From each domination tree, we can read off
a bound on the root of the tree.
Each node is a set of pairs $(z,x)$, where $z$ is a valued object and
$X_z x$.

There are two different types of edge:

(1) Finite disjunction edge.
Partition a domain into finitely many pieces: $X_1,\ldots, X_k$.
Run through each case.

(2) Domination edge.  Pick a subset $S$ of the set at a node, replace
it by $S'$, where $S$ is dominated by $S'$ in the sense of $S'$ having
a greater total value.

The domination trees are combinable: if a subset $S$ in a node is
the root of another domination tree, that tree can be grafted in.

Example 0 (STD CLUSTERS): Root: A centered packing
$(\sigma,\ring{D},D)$. One domination edge. End node: The set of
standard clusters with $\sigma_R(D)$ on each.

Example 1 (REFINED STD REGION):
Root: Any exceptional standard cluster $(D,R)$ with $\sigma$.
One edge:  a domination node.
Set for end-node: $A \cup B \cup C \cup D$.
   \begin{enumerate}
   \item $ A = $ upright and flat quarters in $R$ with $\sigma$
   \item $ B = $ simplices type $C$ in the $S$-system over $R$ with $\sigma(V_S(t_S))$.
   \item $ C = $ little deltas with $-4\doct \vol(\delta_P(v))$.
   \item $ D = $ part of cone of $R$ outside $A,B,C$ with $\sigma_0$.
   \end{enumerate}

Def: Let $\text{anchors} D v$ be the set of anchors in centered
packing $D$ around $\{0,v\}$.

Def: Let $\text{up\_quarter} D v$ be the set of upright quarters in $D$
around $\{0,v\}$.  Let $\text{context} D v$ be the context of $\{0,v\}$.
It is $(p,q)$, where $p = CARD(\text{anchors}D v)$, and
