Sharp Bounds on the Number of Touching Pairs and Triples in a
Packing of Spherical Caps.



This note proves the following assertion, answering a question of
K. Bezdek and S. Reid: the number of touching pairs in an arbitrary
packing of spherical caps of radius $\pi/6$ on $S^2$ is at most $24$.

The proof is obtained by computer search, following the procedure
described in ``A Proof of Fejes Toth's Conjecture on Sphere Packings
with Kissing Number Twelve'' http://arxiv.org/abs/1209.6043.  In fact,
some of the same computer code for that problem can be reused here.
The code has been posted at google code, including graph
generating software, the initial parameters, and the metric coordinate
calculations.  See flyspeck project directory /projects_discrete_geom/bezdek_reid/.

The contact graph is planar and has no loops or multiple joins.  By
increasing the the number of contacts, we may assume that the contact
graph is connected and even $1$-connected.  By arguments in Bezdek and
Reid the degrees are at most $5$.  Each face has at most $8$ sides.
By the kissing number problem, the number of nodes is at most $12$.
In fact, by elementary reasoning (the contact graph must have a
non-triangular face at each node), the number of nodes is $12$.  We
generate all possible graphs of configurations with more than $24$
edges.  By our planar graph generating software, there are $18$ such
graphs, up to isomorphism.

To eliminate the last $18$ graphs we argue as follows. For each case
we form a dual graph, with a vertex for each contact triplet,
connecting vertices by an edge if the original contact triplets share
an edge.  For each connected component of the dual graph, the
coordinates of the spherical caps are determined as soon as the
coordinates of the three spherical caps of a chosen contact triple are
fixed.  A short computer procedure generates the coordinates of
spherical caps separately for each connected component of the dual
graph.  We then impose the following three consistency checks. (1) If
the procedure assigns coordinates to a spherical cap in two ways, the
answers should agree. (2) The angular separation of the spherical caps
must be at least $\pi/3$.  (3) If a pair of spherical caps is
associated with two different connected components of the dual graph,
then their angular separation should be given consistently relative to
the coordinates assigned in the two components.  These three
consistency checks are sufficient to eliminate all $18$ cases.

Similarly, we prove that the number of touching triplets in an
arbitrary packing of spherical caps of radius $\pi/6$ on $S^2$ is at
most $10$.

To prove this, we classify all $1$-connected planar graphs with $12$
nodes at least $11$ triangles.  (Direct geometrical arguments, similar
to those in Bezdek and Reid, rule out the cases with at most $11$
nodes.  Their arguments also reduce us to the case of faces with at
most $8$ sides.)  In this case, we obtain $1335$ graphs up to
isomorphism.  The three consistency checks on the coordinates
described above show that none of these $1335$ has an admissible
metric realization as a contact graph.


Thomas Hales, the author, hereby dedicates this note to the public
domain on January 13, 2013.


