% 
% Author: Thomas C. Hales
% Affiliation: University of Pittsburgh
% email: hales@pitt.edu
%
% latex format

% History.  File started Sep 12, 2011
%% 

\documentclass{llncs}
\pagestyle{headings} 
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{amsmath}

\usepackage{alltt}
\usepackage{rotating}
\usepackage{floatpag}
 \rotfloatpagestyle{empty}
\usepackage{graphicx}
\usepackage{multind}\ProvidesPackage{multind}
\usepackage{times}

% my additions
\usepackage{verbatim}
\usepackage{latexsym}
\usepackage{crop}
\usepackage{txfonts}
\usepackage[hyphens]{url}
\usepackage{setspace}
\usepackage{ellipsis} % 
% http://www.ctan.org/tex-archive/macros/latex/contrib/ellipsis/ellipsis.pdf 

% fonts
\usepackage[mathscr]{euscript} % powerset.
\usepackage{pifont} %ding
\usepackage[displaymath]{lineno}
\usepackage{rotating}

% automatically generate revision number by
% svn propset svn:keywords "LastChangedRevision" sdodec.tex
\def\svninfo{{\tt
  filename: sdodec.tex\hfill\break
  PDF generated from LaTeX sources on \today; \hfill\break
  Repository Root: https://flyspeck.googlecode.com/svn \hfill\break
  SVN $LastChangedRevision$
  }
  }
%-%

% Math notation.
\def\op#1{{\hbox{#1}}} 
\def\tc{\hbox{:}}
%\newcommand{\ring}[1]{\mathbb{#1}}
\def\amp{\text{\&}}
\def\bq{\text{\tt {`}\,}}
\def\true{\text{true}}
\def\false{\text{false}}
\def\princ#1{\smallskip\hfill\break\smallskip\centerline{\it #1\hfill}}
% Flags

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\title{The Strong Dodecahedral Conjecture and\\ Fejes T\'oth's Contact Conjecture}
\author{Thomas C. Hales\thanks{{Research supported in part by 
NSF grant 0804189 and the Benter Foundation.}}}
\institute{University of Pittsburgh\\
\email{hales@pitt.edu}}
\maketitle


\begin{abstract} This article sketches the proof of two theorems in
  the theory of sphere packings in Euclidean $3$-space.  The first is
  K. Bezdek's strong dodecahedral conjecture: the surface area of
  every bounded Voronoi cell in a packing of balls of radius $1$ is at least
  that of a regular dodecahedron of inradius $1$.  The second theorem
  is L. Fejes T\'oth's contact conjecture, which asserts that in
  $3$-space, any packing of congruent balls such that each ball is
  touched by twelve others consists of hexagonal layers.  Both proofs
  are computer assisted.  Complete proofs of these theorems
  appear in the author's forthcoming book ``Dense Sphere Packings.''
\end{abstract}

\input {../../kepler_tex/kepmacros.tex}

\section{The strong hexagonal conjecture}

To describe methods, we begin with a  proof of the following theorem in $\ring{R}^2$.  

\begin{theorem}  The perimeter of any bounded Voronoi cell of a packing of congruent balls of radius $1$
in $\ring{R}^2$ is at least $4\sqrt{3}$, the perimeter of a regular hexagon with inradius $1$.
\end{theorem}

If we adopt the convention that the perimeter of an unbounded Voronoi cell is infinite, then
the boundedness hypothesis can be dropped from the theorem.

\begin{proof} Fix a bounded Voronoi cell of in a packing of congruent
  balls of radius $1$.  We fix a coordinate system with the center of
  the Voronoi cell at the origin.  The intersection of the Voronoi
  cell with a disk of radius $\sqrt2$ at the origin is convex disk
  whose boundary $C$ consists of circular arcs and straight line
  segments.  The length of $C$ is no greater than the original
  perimeter of the Voronoi cell.  It suffices to show that the length
  of $C$ is at least $4\sqrt{3}$.



The boundary $C$ is determined by the set of centers $V$ of balls at
distance less than $\sqrt8$ from the origin, excluding the ball
centered at the origin.

The following function arises in the proof of the dodecahedral
conjecture in three-dimensions.  Its purpose will be described later.
\begin{equation}
L(h) = \begin{cases} 
   (h_0-h)/(h_0-1),& \text{if } h \le h_0,\\
  0,&\text{ otherwise},
 \end{cases}
\end{equation}
where $h_0 = 1.26$.  

Let $\u_1,\u_2\in V$ be distinct points such that
$T=\{\orz,\u_1,\u_2\}$ has circumradius less than $\sqrt2$.  Let
$\ell(\u_1,\u_2)$ be the part of $C$ contained in the convex hull of
$T$, and let $\theta$ be the angle at $\orz$ between $\u_1$ and
$\u_2$.  We look for constants $b,c\in\ring{R}$ such that the
following conditions hold:
\begin{equation}\label{eqn:ineq}
\ell(\u_1,\u_2) - (b \theta(\u_1,\u_2) + c (L(\normo{\u_1)}/2) + L(\normo{\u_2}/2))) \ge 0,
\end{equation}
and equality holds when $T$ is an equilateral triangle with side $2$.

In this, the planar case, the regular hexagon is a {\it corner
  solution} with $\norm{\u_1}{\u_2}=2$.  In the three-dimensional
case, the regular dodecahedron will satisfy $\norm{\u_1}{\u_2}>2$.  It
is natural to require $b,c$ to be chosen so that the left-hand side of
(\ref{eqn:ineq}) has a local minimum at $t=2$ along the curve \dots.


We form a graph $(V,E)$ by connecting two points $\u,\v$
 of $V$
by an edge if the circumradius of $\orz,\u,\v$ is less than $\sqrt2$.  
\end{proof}

\section{Marchal cells}

\section{Strong Dodecahedral Conjecture}

\section{Fejes T\'oth's Contact Conjecture}

\raggedright
\bibliographystyle{amsalpha} % was plain %plainnat
\bibliography{/Users/thomashales/Desktop/googlecode/flyspeck/latex/bibliography/all}


\bigskip
\noindent
\svninfo
\smallskip

\smallskip
(Draft: do not circulate) % XX





\end{document}

